### 2012-08-02 ResponseTo_GW20120518-0099_Review.tex
Original commit

\documentclass[12pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

\usepackage{geometry} % to change the page dimensions
\geometry{letterpaper} % or letterpaper (US) or a5paper or....
\geometry{margin=1in} % for example, change the margins to 2 inches all round

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{color}

\usepackage{amsmath}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim

\begin{document}

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
\textbf{Add some introductory text about how reviewer comments were helpful and have been addressed in the revised manuscript}

Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.

\section*{Editor's Comments:}
\begin{enumerate}
\item \textit{Our layout doesn't require section numbering, so please delete those.}

Section numbering has been eliminated in the revised manuscript.

\end{enumerate}

\section*{Reviewer 1 Comments:}
\begin{enumerate}
\item \textit{The introduction should also introduce the speedup potential and bottlenecks of GPGPU and parallel CPU processing for the UPCG solver.}

Response.

\item \textit{Section 2 has too much jargon. I suggest removing what is irrelevant and explaining what is, in terms of how that helps speed up or slow down GPGPU processing compared to serial computing or multi-core processing on CPUs.}

Response.

\item \textit{The manuscript needs to mention how the solver is available for M2K5 for the GPGPU and multi-core CPU versions (are they DLLs, how are they linked into the main code, etc). I do not expect people to generally compile their codes, let alone within CUDA. Some direction therefore is needed on accessibility.}

A section titled ``How to Obtain the Software'' has been added that provides information on how to obtain the software.

\item \textit{Also, direction is needed on hardware that may be best suited for the GPUGPU processing will benefit this manuscript.}

The authors are not experts on the hardware that is best suited for GPGPU processing. Rather than providing guidance on appropriate hardware we feel it is more appropriate for potential users to consult with a qualified software dealer and identify the hardware that is most appropriate for their particular organization.

\item \textit{GPGPU speedup has been compared to the PCG5 solver on a serial CPU. With more advanced solvers coming into the MODFLOW family (for example the XMD solver in MODFLOW-NWT,)it would be interesting to note the speedup of UPCG compared to even more robust higher level ILU solvers which may be even faster than PCG5 for hard problems.}

The analyses presented in this manuscript are not meant to be an exhaustive comparison of GPGPU solutions but an evaluation of the speedup possible.

\item \textit{Issues with GPGPU processing for different schemes are scattered throughout the manuscript. These issues should be collated into a separate section or a table. For instance: GPGPU memory requirements for larger problems; need to maximize the amount of time spent in the linear solvers; the overhead associated with GPGPU-CPU memory copy operations relative to the solution time; higher percentage of runtime spent transferring preconditioner data; why the total time spent applying the preconditioner actually increases for the MILU0 preconditioner; what is less optimized in Intel FORTRAN for multi-core processing; details on problem size limitation due to RAM available on the GPGPU; and others.}

Response.

\end{enumerate}

\section*{Reviewer 2 Comments:}
\begin{enumerate}
\item \textit{Overall, lots of jargon is used. Naturally, this is an emerging field, and one that many readers of Ground Water will not be familiar with, so the use of jargon will be difficult for the uninitiated reader. For example, lines 75-79 are filled with terms that will mean nothing to the average groundwater modeler - consider deleting (or moving to appendix or online material) this paragraph and other similar paragraphs that might not hold much meaning to the average reader.}

Response.

\item \textit{Thorough comparisons were made for different preconditioners and different parallelization strategies and the authors did a great job of making the comparisons fair. Considering the results of the various preconditioners, it appears that Jacobi preconditioning is good enough for many of the problems, which leads me to believe that the K fields are fairly smooth (and the spread of the eigenvalues are nicely clustered). I tend to think of this as the exception in rather than the rule in most practical groundwater models. Can you provide guidance for using MIC vs. POLY vs. Jacobi based on the K field? In other words, I think these results are highly dependent on the K field and only 2 fields are used. How would the results vary if you stretched the contrasts in your K field? In that regard, to be fair, the GMG solver should also be used as a basis for comparison and give insight about which strategy should be used for optimal performance.}

Response.

\item \textit{There seems to be a tradeoff between parallelizing on the CPU vs. the GPU depending on the sophistication of the preconditioner. What is not clear was if it Open-MP was used on PCG2. Reference is made to the work by to Dong and Li, but it appears that this was not considered. How do these results compare to that of Dong and Li? The intent here is not to say that your results are better/worse than those of Dong and Li. Simply, not everyone has access to a video card with double precision GPU capabilities, whereas multicore CPU's seem to be more common, so some guidance of which strategy might be more effective depending on the user's hardware would be helpful (ie, If I were buying a new workstation, should I focus on getting the best multicore CPU, or should I get a less powerful CPU and spend money on a high-end GPU?) I understand that the answer to this question is not straightforward, but any additional insights you could provide regarding this tradeoff would be helpful.}

Response.

\item \textit{Line 26 - "can be *easily* parallelize"}

Response.

\item \textit{Lines 28 - 34 - there is some time discrepancies here from 2005 to 1984, to 1988. Perhaps just say "MODFLOW" on line 28 and remove the "since 1984" and insert "numerous" before "two-and" }

Response.

\item \textit{Line 47 "present" (plural)}

Response.

\item \textit{Line 48 - unclear - faster at what exactly? Operations,cycles/second? }

Response.

\item \textit{Lines 75-108 is filled with a lot of technical jargon. Consider moving some of this to an appendix (as noted above). }

Response.

\item \textit{Line 108 - really, the parallelization you mention can be applied to many simulators, not just MODFLOW-2005. Consider mentioning wider range applicability. }

Response.

\item \textit{Line 213-214 - unclear. Consider rephrasing }

Response.

\item \textit{Line 351 - "developed to *use* the PCG" (one does not "solve" the algorithm) }

Response.

\item \textit{Line376 - as indicated above, can you provide some guidance here about when one approach might be better than another depending on the system properties (size, K field). Also, GMG should be included here to so that the full suite of approaches are compared (weak but highly parallelizable to strong but not very parallelizable). }

Response.

\item \textit{Figure 4 - I think this figure is extremely important as it shows where the speed ups occur. However, it is not clear how much time is spent in the GPU for the matrix-vector and BLAS operations because the bars seem to not show up on the scale? Can you include a value here so the reader does not assume these are zero. }

Response.

\item \textit{In light of the speed ups coming from CUBLAS operations being parallelized over the cells in the model, I am wondering if additional speed up could be obtained by using other primitives in the CUSparse library for the triangular solves and/or the decomposition? Was this considered as well? There is some new work that tries to parallelize ILU (see: http://research.nvidia.com/publication/parallel-incomplete-lu-and-cholesky-factorization-preconditioned-iterative-methods-gpu) }

Response.

\item \textit{As indicated above lines 75-108 are filled with jargon and should be reconsidered. 
Response.}

Response.

\end{enumerate}

\section*{Reviewer 3 Comments:}
\begin{enumerate}
\item \textit{Quality of figures should be improved (all figures but in particular, figures 3 to 6). Texts (titles and legends) are too small to read. Lines are very hard to read.}

Response.

\item \textit{Tables should be reworked. I do not think we need 3 digits after decimal point for these comparisons. Please also consider significant digits in the tables.}

Response.

\item \textit{The authors only consider total execution time which includes time required for both Picard and linear solver iterations. Did you closely examined these iteration performance in solver comparisons? How these iteration performance changes among different preconditioning schemes? }

Response.

\item \textit{Lines 28-32: the authors mentioned computational performance changes due to domain discretization. Does this cause by FD formulation or matrix solver scheme? Explanation is needed. }

Response.

\item \textit{Line 48: 448 $\rightarrow$ 448 cores. }

Response.

\item \textit{Line 114: is L3/L3T correct? }

Response.

\item \textit{Line 123: is L2/T correct? }

Response.

\item \textit{Line 161: What is CUBLAS? }

Response.

\item \textit{Line 199: Problem Description. Reasoning for choosing test cases is helpful for the readers. }

Response.

\item \textit{Line 236: "(10-0)/1000"is not necessary.}

Response.

\end{enumerate}

\end{document}

### 2012-08-02 GW20120518-0099.tex
Original commit

\documentclass[12pt]{article} 
\usepackage[utf8]{inputenc}

%%% PAGE DIMENSIONS
\usepackage{geometry}
\usepackage{lineno}
\usepackage{color}
\usepackage[kernelfbox]{xcolor}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
%%% LANGUAGE HYPHENATION
\usepackage[english]{babel}
%%% HEADERS & FOOTERS
\usepackage{fancyhdr} 
%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
%%% MATHS RELATED PACKAGES
\usepackage{import}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{setspace}
\usepackage{array}
\usepackage{multirow}
%%% OTHER PACKAGES
\usepackage{booktabs}
\usepackage{array}
\usepackage{paralist}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{url}

%%% SETTINGS
\geometry{letterpaper}
\geometry{margin=1in}
\doublespacing
\linenumbers
\allsectionsfont{\sffamily\mdseries\upshape} 
\graphicspath{{./Figures/}}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\lhead{}\chead{Hughes and White -- MODFLOW-2005 UPCG GPGPU Solver}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

\title{Use of general purpose graphical processing units with MODFLOW-2005}
\author{Hughes, J.D.\\U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jdhughes@usgs.gov} \and White, J.T. \\ U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jwhite@usgs.gov}}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
\begin{abstract}
To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW-2005, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed. The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete LU factorization, and generalized least-squares polynomial preconditioners. The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP. For simulations utilizing the GPGPU, all basic linear algebra operations are performed on the GPGPU; memory copies between the central processing unit CPU and GPCPU occur prior to the first iteration of the UPCG solver and after satisfying head and flow criteria or exceeding a maximum number of iterations. The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells. Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW-2005 PCG solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize. Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be parallelized.
\end{abstract}

\section*{Introduction}

MODFLOW-2005 (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to two- and three-dimensional problems since 1984. Because MODFLOW-2005 uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively. To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers. The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (SOR) linear solver. These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG2) linear solver (Hill 1990), the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication bottlenecks which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.

\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs. Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7. In addition to OpenMP and MPI, GPGPUs also presents an option for linear solver parallelization. In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs). Use of GPGPUs to parallelize linear solvers for MODFLOW-2005 is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation. For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems. For solution of the groundwater flow equation, Ji \textit{et al.} (2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.

In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured UPCG linear solver for a unconfined groundwater flow problem. We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete LU (MILU0), and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU. And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.

\section* {\bf GPGPU Parallel Programming}
GPGPU code developed using the NVIDIA$^{\textregistered}$ CUDA API is essentially a sequential code that is capable of using the fork-join model of parallel execution similar to OpenMP. In a fork-join model, a single master thread is active when program execution begins. The master thread executes sequential portions of the code. At points where parallel operations are required, the master thread spawns, or forks, additional threads that work concurrently with the master thread through the parallel section. At the end of the parallel code, the spawned threads are suspended and rejoined to the master thread. The fork-join model used in OpenMP is shown graphically in \textbf{figure~\ref{FigParallel}}. Typically, the number of OpenMP threads used in the fork-join model is less than the total number of cores available on multi-core CPUs.

For GPGPU code implementations, the concept is similar to OpenMP.  The main code is a sequential host code which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.  \color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have to potential to spawn thousands to tens-of-thousands of threads to execute the CUDA kernel function (\textbf{Box 2}). The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads. \color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} not just for MODFLOW-2005, but for most numerical modeling applications\color{black}.

\section* {\bf Conjugate Gradient Linear Solver}
The constant-density, three-dimensional groundwater flow equation is described by the partial-differential equation,

\begin{equation}
	\label{gwpde}
	\frac{\partial}{\partial x} \left( K_{xx} \frac{\partial h}{\partial x} \right) + 
	\frac{\partial}{\partial y} \left( K_{yy} \frac{\partial h}{\partial y} \right) + 
	\frac{\partial}{\partial z} \left( K_{zz} \frac{\partial h}{\partial z} \right) + 
	W = S_{S} \frac{\partial h}{\partial t} ,
\end{equation}

\noindent
where  $K_{xx}$, $K_{yy}$, $K_{zz}$ are the hydraulic conductivity [L/T] along the x, y, and z-coordinate axes which are assumed to be parallel to the major axes of hydraulic conductivity; $h$ is the groundwater head [L]; $W$ is a volumetric flux per unit volume [L$^{3}$/L$^{3}$T] representing sources/sinks of water; S$_{S}$ is the specific storage [L$^{-1}$]; and t is time [T].

MODFLOW-2005 uses a cell-centered, finite-difference approximation of \textbf{equation~\ref{gwpde}} to solve for three-dimensional groundwater flow (Harbaugh 2005). Development of the finite-difference form of \textbf{equation~\ref{gwpde}} using the continuity equation for a cell results in

\begin{equation}
	\label{conteq}
	\sum Q = S_{S} \frac{\Delta h}{\Delta t} V,
\end{equation}

\noindent
where $Q$ is the flow rate [L$^{3}$/T] into a cell from adjacent cells and sink/source terms, $\Delta h$ is the head change [L] in a cell over a time interval $\Delta t$ [T], and $V$ is the volume of a cell [L$^{3}$]. Manipulation of \textbf{equation~\ref{conteq}} to separate known and unknown $h$ terms results in a large system of simultaneous linear equations of the form,

\begin{equation}
	\label{lineareq}
	\mathbf{Ah} = \mathbf{b},
\end{equation}

\noindent
where $\mathbf{A}$ is a  square, symmetric, positive-definite coefficient matrix [L$^{2}$/T] that includes cell-by-cell conductances, calculated using cell dimensions and hydraulic conductivities at cell interfaces, and unknown components of sink/source and storage terms; $\mathbf{h}$ is the vector of unknown heads at time $t$; and $\mathbf{b}$ is a known vector of source/sink and storage terms.

\subsection*{Conjugate Gradient Algorithm}

The conjugate gradient (CG) iterative method is an efficient method for solving large systems of linear equations having a square, symmetric, positive-definite coefficient matrix. Pseudocode for the preconditioned conjugate gradient method is given in \textbf{algorithm~\ref{pcgalg}}.

\begin{algorithm}
\caption{Preconditioned Conjugate Gradient Method}
\label{pcgalg}
\begin{algorithmic}[1]
	\State $\text{Compute } \mathbf{r}_{0} = \mathbf{b - Ah}_{0} \text{ for some initial guess } \mathbf{h}_{0}$
	\For {$i = 1 \to maxinner$}
		\State $\textbf{solve } \mathbf{Mz}_{i-1} = \mathbf{r}_{i-1}$ \Comment{apply preconditioner}
		\State $\rho_{i-1} = \mathbf{r}^T_{i-1} \mathbf{z}_{i-1}$
		\If {$i = 1$}
			\State $\mathbf{p}_i = \mathbf{z}_0$
		\Else
			\State $\beta_{i-1} = \rho_{i-1} / \rho_{i-2}$
			\State $\mathbf{p}_i = \mathbf{z}_{i-1} + \beta_{i-1} \mathbf{p}_{i-1}$
		\EndIf
		\State $\mathbf{q}_i = \mathbf{Ap}_{i}$
		\State $\alpha_i = \rho_{i-1} / \mathbf{p}^T_i \mathbf{q}_{i}$ \Comment{calculate step-size}
		\State $\mathbf{h}_i = \mathbf{h}_{i-1} + \alpha_i \mathbf{p}_{i}$  \Comment{update head}
		\State $\mathbf{r}_i = \mathbf{r}_{i-1} - \alpha_i \mathbf{q}_{i}$ \Comment{update residual}
		\State $\text{check convergence; continue if necessary}$
	\EndFor
\end{algorithmic}
\end{algorithm}

In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the residual of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless].

All linear iteration steps, except for application of the zero fill-in incomplete LU (ILU0) and MILU0 preconditioners in line 3 of \textbf{algorithm~\ref{pcgalg}}, include operations that are independent for each active groundwater cell, indicating a potential for parallelization. In cases where (1) no preconditioner ($\mathbf{z = r}$) or (2) a Jacobi preconditioner ($\mathbf{M = I D}^{-1}$ -- where $\mathbf{I}$ is the identity matrix and $\mathbf{D}^{-1}$ is a matrix that only contains the inverse of the diagonal elements of $\mathbf{A}$), is selected, application of the preconditioner is highly parallelizable. For most practical problems, non-preconditioned and the Jacobi preconditioners will not significantly reduce the number of linear iterations ($i$) required to achieve convergence (Li and Saad 2010). As a result better preconditioners, such as incomplete factorizations (ILU0 or MILU0) or m-degree polynomial (GLSPOLY) approximations of $\mathbf{A}$, are generally needed. Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994). Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize (Liang \textit{et al.} 2002; Saad 2003).

\subsection*{Numerical Implementation}
The UPCG linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations. Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.

Although the connectivity of the finite-difference discretization used in MODFLOW-2005 uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the CUBLAS libraries. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Sadd 2003).

Available CUBLAS matrix-vector products and level-1 BLAS operations were used to parallelize the GPGPU capabilities of the UPCG linear solver. Because of the poor performance of forward and back substitution operations used by the ILU0 and MILU0 preconditions observed during initial testing of the UPCG linear solver and observed by others (Li and Saad 2010), application of these preconditioners is performed on the CPU, resulting in a hybrid CPU/GPGPU solver. The results of the matrix-vector product, calculated by application of the ILU0 and MILU0 preconditioners on the CPU, is accessed as page locked memory by the GPGPU to achieve higher memory bandwidth.

For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and level-1 BLAS operations. The OpenMP matrix-vector product was constructed using a level-2 BLAS GEMV approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004). OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively. The number of OpenMP threads used for matrix-vector products and level-1 BLAS operations is specified separately in the UPCG input file. For sequential CPU simulations,  separate matrix-vector product and level-1 BLAS operation routines that exclude OpenMP directives are used to eliminate all OpenMP overhead.

Infinity norms ($\left \|  x \right \|_{\infty}$) are used to evaluate convergence of the UPCG linear solver with respect to the change in $\mathbf{h}$ (HCLOSE) and $\mathbf{r}$ (RCLOSE). In MODFLOW-2005, non-linearities are resolved using Picard iteration. Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver. This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of outer (Picard) iterations is exceeded. \\

\section* {\bf Test Cases}
The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing perfomance (NVIDIA compute capability 2.0). The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel Xenon 3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS. For comparison with GPGPU results, parallel CPU simulations were executed on dual 4-core Intel Xenon 2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).

\subsection*{Problem Description}
GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a 1,000 m square heterogeneous unconfined aquifer.  The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively. Models with horizontal discretizations of 200 $\times$ 200, 500 $\times$ 500, 1,000 $\times$ 1,000, 2,000 $\times$ 2,000, and 4,000 $\times$ 4,000 cells per layer were evaluated. Single layer models were evaluated for all horizontal discretizations and multi-layer models were evaluated for select horizontal discretizations to determine the effect that a 7-point stencil, and the resulting additional non-zero entries for each cell, has on GPGPU and parallel CPU execution time. Two- and three-layer models were evaluated for all horizontal discretizations except 4,000 $\times$ 4,000 cells per layer. Ten-layer models were also evaluated for horizontal discretizations ranging from 200 $\times$ 200 to 1,000 $\times$ 1,000 cells per layer. Additional horizontal and vertical discretizations were not evaluated because of GPGPU memory requirements for larger problems. For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers. The thickness of layer 1 was calculated to be sum of the top elevation of the model (10 m) and constant thickness used for layers 2 and below.

The heterogeneous hydraulic conductivity distribution (base hydraulic conductivity data) was generated for the 1,000 $\times$ 1,000 cell model (base model) using sequential Gaussian simulation with log-transformed (base 10) parameters based on a mean value of 5 m/d and an exponential variogram having a nugget of 1.023 $\log_{10}((m/d)^{2})$, a variance of 1.23 $\log_{10}((m/d)^{2})$, and length parameter ($a$) of 250.5 m (effective range $\approx$ 750 m). The axis of anisotropy was rotated by 45 degrees to the model grid and a 3.5:1 ratio of horizontal anisotropy was used. For the realization selected and used to evaluate GPGPU and parallel CPU results, the minimum, average, and maximum hydraulic conductivity in the model domain were 0.43, 4.81, and 43.3 m/d, respectively. The distribution of hydraulic conductivity in the model domain is shown in \textbf{figure~\ref{FigModelDomain}}.

Hydraulic conductivity was not varied between layers in the multi-layer models evaluated. For models having less than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was averaged for each base model cell contained in a coarse model cell. For models having more than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was directly assigned to each higher-resolution cell contained within a coarse base model cell.

Head-dependent (general head) boundaries were specified on the left and right sides and no flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain $\left [ \left(10. - 0.0\right) / 1000. = 0.01 \text{ m/m} \right]$ and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

For all model discretizations, the UPCG solver with the Jacobi, MILU0, and GLSPOLY preconditioners were evaluated on the CPU, using both the sequential and parallel (OpenMP) options, and GPGPU. To maximize the amount of time spent in the linear solvers the maximum number of outer (Picard) and inner (linear) iterations were set to 50 and 1,000, respectively. A HCLOSE value of 0.001 m was used for all simulations. In order to use a consistent RCLOSE value in all simulations, RCLOSE was calculated as the product of HCLOSE and the cell area, effectively scaling RCLOSE with the discretization. A degree 10 polymonial was constructed for simulations using the GLSPOLY preconditioner and because the coefficient matrix $\mathbf{A}$ was scaled using diagonal scaling we assumed the minimum and maximum eigenvalues for the scaled matrix were between 0.0 and 2.0 (Scandrett 1989).

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.

\subsection*{Results}
The CPU/GPGPU speedup of UPCG solver simulations for the Jacobi, MILU0, and GLSPOLY preconditioners are summarized on \textbf{table~\ref{TableGPUCPUSpeedup}}. Significant  CPU/GPGPU speedups, ranging from 1.5 to 35, are observed with Jacobi and GLSPOLY preconditioners. Only marginal speedups, ranging from less than 1.0 to almost 2.0, are observed with the MILU0 preconditioner. The maximum number of outer (Picard) and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems. The smallest CPU/GPGPU speedup corresponds with the smallest model discretizations (200 $\times$ 200 $\times$ 1) because of the overhead associated with GPGPU-CPU memory copy operations relative to the solution time.

The CPU/GPGPU speedup of MILU0 and GLSPOLY simulations are shown graphically on \textbf{figure~\ref{FigCPUGPUResults}}. The CPU/GPGPU speedup of MILU0 simulations are approximately 2 for simulations with more than 3 million active model cells. The reduction in the CPU/GPGPU speedup of MILU0 simulations between 1 and 3 million active model cells is a result of the higher percentage of runtime spent transferring preconditioner data ($\mathbf{r \text{ and } z}$ in \textbf{algorithm~\ref{pcgalg}}) between the GPGPU and CPU during each linear iteration for application of the MILU0 preconditioner. The CPU/GPGPU speedup of  GLSPOLY preconditioner increases to approximately 30 for simulations with more than 5 million active cells; the increase in the CPU/GPGPU speedup with increased problem size for the Jacobi preconditioner is comparable to the GLSPOLY preconditioner (\textbf{fig.~\ref{FigCPUGPUResults}b}).

The amount of time applying the MILU0 and GLSPOLY preconditioners, performing matrix-vector operations, level-1 BLAS operations, and transferring data between the CPU and GPU for a select number of simulations are shown graphically in \textbf{figure~\ref{FigBLASOperationResults}}. Matrix-vector operations account for approximately half of the time spent in the UPCG solver for sequential CPU simulations using either the MILU0 or GLSPOLY preconditioners. On the GPGPU, matrix-vector operations for the MILU0 and GLSPOLY preconditioner are significantly reduced, largely as a result of the ability of the GPGPU to apply massively parallel resources to perform BLAS operations.  Analysis of operation execution times indicates that the MILU0 speedup is notably less than the GLSPOLY speedup because the time spent applying the MILU0 is essentially the same whether the GPGPU is used or not.  This is because the application of the MILU0 preconditioner is always performed on the CPU.  For the MILU0 preconditioner, all of the speedup occurs during the conjugate gradient routine in the level-1 BLAS operations [matrix-vector multiplication and vector operations (dot products)], which are carried out on the GPGPU after application of the preconditioner. Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases. For the GLSPOLY preconditioner, solution on the GPGPU reduces the time spent completing linear solver operations by more than 95\% and the time spent transferring data between the CPU and GPGPU is negligible.

\section* {\bf Discussion}
Although notable, CPU/GPGPU speedups summarized on \textbf{table~\ref{TableGPUCPUSpeedup}} are not a fair measure of the speedups that would result from use of the GPGPU to solve a MODFLOW-2005 problem. A more realistic comparison is the speedup of GPGPU simulations relative to sequential CPU simulations using the standard MODFLOW-2005 PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990). PCG(MIC)/GPGPU speedups for the Jacobi, MILU0, and GLSPOLY preconditioners are summarized on \textbf{table~\ref{TableGPUSpeedup}}.  PCG(MIC)/GPGPU speedups for the MILU0 and GLSPOLY preconditioners are shown graphically on \textbf{figure~\ref{FigGPUResults}}.

PCG(MIC)/GPGPU speedups for the MILU0 preconditioner are comparable to CPU/ GPGPU speedups; the similarity of PCG(MIC)/GPGPU and CPU/GPGPU speedups for the MILU0 preconditioner are a result of the similarity in time required to perform forward and backward substitutions during application of the preconditioner with the PCG and UPCG solvers, as the MIC is similar to MILU0 . PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners range from less than 1.0 to more than 8.0, which are notably less than CPU/GPGPU speedups.  This is attributable to the efficiency of the standard MODFLOW-2005 PCG solver with the MIC preconditioner. The relative increase in PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).

Parallel CPU simulations were run to compare speedups possible with GPGPUs to speedups that could be realized on multi-core CPUs using the OpenMP programming paradigm. PCG(MIC)/UPCG-OpenMP speedups for the Jacobi, MILU0, and GLSPOLY preconditioners using 4, 7, and 14 OpenMP threads are summarized on \textbf{table~\ref{TableOpenMPSpeedup}}. PCG(MIC)/UPCG-OpenMP speedups for the Jacobi and GLSPOLY preconditioners are notably less than PCG(MIC)/GPGPU speedups and is expected because the number of threads on the GPGPU is orders of magnitude greater than for parallel CPU simulations. For the MILU0 preconditioner, PCG(MIC)/UPCG-OpenMP speedups (average speedup $=$ 1.539) are comparable to PCG(MIC)/GPGPU speedups (average speedup $=$ 1.447) and there is little benefit to using the GPGPU with the MILU0 preconditioner. PCG(MIC)/UPCG-OpenMP speedups for the MILU0 preconditioner with 7 OpenMP threads are shown graphically on \textbf{figure~\ref{FigOpenMPResults}}. It should be noted that the Intel Fortran compiler (version 12.1.2.278), with the highest-level optimization option (/O3 compiler switch), was used to compile MODFLOW-2005 with the UPCG solver. The relatively small PCG(MIC)/UPCG-OpenMP speedups observed with the UPCG solver compiled using the Intel Fortran compiler is consistent with the observations of Dong and Li (2009). Larger PCG(MIC)/UPCG-OpenMP speedups would likely be observed with other compilers but the value of using a less optimized compiler is questionable.

\section* {\bf Conclusions}

The PCG algorithm was implemented for use on GPGPUs using the NVIDIA CUDA-enabled implementation of the BLAS library (CUBLAS). The availability of this API makes it relatively simple to develop parallel GPGPU code. The fork-join parallel model, which uses multiple threads in parallel sections, is used to parallelize code on GPGPUs. Unlike fork-join approaches used on CPUs (OpenMP), which have thread limits determined by the number of CPUs and cores per CPU, GPGPUs have the capability of using thousands to tens of thousands of threads concurrently in parallel sections. The number of concurrent threads that can be used on GPGPUs makes them ideal for parallelization of simultaneous solutions of systems of linear equations like those solved by MODFLOW-2005. The key to successful parallelization on GPGPUs is to (1) use numerical approaches that are highly parallelizable and (2) reduce the need for memory transfers between the CPU and GPGPU as much as possible.

The UPCG solver was developed to solve the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners. An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library. The UPCG solver also includes options for running sequential and parallel simulations on the CPU. Parallel CPU simulations are executed using the OpenMP programming paradigm.

A number of test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW-2005.  CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared to CPU runtimes for the same preconditioner. Exceptions to notable CPU/GPGPU speedups were observed for the smallest problems where CPU-GPGPU memory transfer times were significant compared to solution time. CPU/GPGPU speedups with the MILU0 preconditioner was the smallest of all of the preconditioners (ranging from 0.99 to 1.936) because of the additional CPU-GPGPU memory transfers required during each inner (linear) iteration to apply the preconditioner.

Speedups calculated using GPGPU runtimes and CPU runtimes for the same solver are not fair measures of the benefits of using GPGPUs to solve the groundwater flow equation because of the efficiency of the standard MODFLOW-2005 PCG solver with the MIC preconditioner. PCG(MIC)/GPGPU speedups are smaller than speedups calculated using sequential CPU runtimes for the UPCG solver. Use of GPGPUs with the UPCG solver and the GLSPOLY preconditioner can result in PCG(MIC)/GPGPU speedups as large as 8 for large multi-layer models when compared to the standard MODFLOW-2005 PCG solver with the MIC preconditioner. For certain problems, the Jacobi preconditioner is adequate and has similar PCG(MIC)/GPGPU speedups.

When compared to parallel CPU solutions, speedups for GPGPU solutions using the UPCG solver with the Jacobi and GLSPOLY preconditioners are significantly better. For problems where the MILU0 preconditioner is needed there is no real benefit to using the GPGPU because of the required additional CPU-GPGPU memory transfers -- for these cases parallel CPU solution would be adequate.

In summary, use of GPGPU results in significant speedups when the Jacobi and GLSPOLY preconditioners are capable of reducing the total number of inner (linear) iterations required to achieve specified convergence criteria. The benefits of using the GPGPU increases as the problem size increases. In this study, problem size was limited by the RAM available on the GPGPU. Larger problems could be solved with GPGPUs having more RAM or extending the UPCG solver to allow use of multiple GPGPUs. The UPCG solver is designed to solve  square, symmetric, positive-definite coefficient matrices. Modification of the UPCG solver to allow solution of nonsymmetric matrices, using the BiConjugate Gradient Stabilized method or other similar method (Barrett \textit{et al.} 1994), would permit use of the UPCG solver with MODFLOW-NWT (Niswonger \textit{et al.} 2011), MT3DMS (Zheng and Wang 1999), or SEAWAT (Langevin \textit{et al.} 2007).

\section* {\bf Acknowledgments}

The authors are grateful to Vevek Bedekar, Randall T. Hanson, and Yong Liu for the valuable suggestions and helpful comments on previous versions of this article. The authors are also grateful to the editor and three anonymous reviewers for their valuable suggestions and comments.

\color{blue}
\section* {\bf How to Obtain the Software}

Source code for MODFLOW--2005 with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}. Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit which includes the necessary CUDA drivers. NVIDIA also provides GPGPU code samples in the CUDA software development kit (SDK). Current versions of each of these are available free of charge from NVIDIA at \url{http://www.nvidia.com/content/cuda/cuda-downloads.html}. A 64-bit version of the MODFLOW--2005 executable with the UPCG solver and the linked OpenMP dynamic-link library (DLL) for the Windows 7 operating system and the Tesla C2050 GPGPU is also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg} and only requires installation of the CUDA drivers, which are also available as a stand-alone installation from NVIDIA, and a GPGPU with NVIDIA Compute Capability 2.0 or greater. This executable was linked against version 4.1.28 of the 64-bit CUDA toolkit. It is possible that the software may need to be modified for application of the code on GPGPU models other than the Tesla C2050. And finally, python scripts used to create the MODFLOW-2005 datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.

The USGS provides no warranty, expressed or implied, as to the correctness of the furnished software or the suitability for any purpose. The software has been tested, but as with any complex software, there could be undetected errors. Users who find errors are requested to report them to the USGS. References to non-USGS products, trade names, and (or) services are provided for information purposes only and do not constitute endorsement or warranty, express or implied, by the USGS, U.S. Department of Interior, or U.S. Government, as to their suitability, content, usefulness, functioning, completeness, or accuracy.\color{black}

\section*{\bf References}
\begin{description}

\item Barrett R., Berry, M.,  Chan, T.F., Demmel, J., Donato, J., Dongarra, J., Eijkhout, V.,  Pozo, R., Romine, C., Van Der Vorst, H. 1994. Templates for the Solution of Linear Systems, Building Blocks for Iterative Methods. Philadelphia, Pennsylvania, SIAM, 124 p.

\item Dong, Yanhui, Li, Guomin 2009. A Parallel PCG Solver for MODFLOW. Ground Water 47: 845–850. doi: 10.1111/j.1745-6584.2009.00598.x

\item Harbaugh, A.W. 2005. MODFLOW-2005, the U.S. Geological Survey modular ground-water model -- the Ground-Water Flow Process. U.S. Geological Survey Techniques and Methods 6-A16, variously p.

\item Hill, M.C. 1990. Preconditioned conjugate-gradient 2 (PCG2), a computer program for solving ground-water flow equations. U.S. Geological Survey Water-Resources Investigations Report 90-4048, 43 p.

\item Ji, Xiaohui, Cheng, Tangpei, Wang, Qun 2012. CUDA-based solver for large-scale groundwater flow simulation. Engineering with Computers 28, no. 1: 13-19. doi: 10.1007/s00366-011-0213-2

\item Langevin, C.D., Thorne, D.T., Jr., Dausman, A.M., Sukop, M.C., and Guo, Weixing 2007. SEAWAT Version 4: A Computer Program for Simulation of Multi-Species Solute and Heat Transport. U.S. Geological Survey Techniques and Methods Book 6, Chapter A22, 39 p.

\item Li, Ruipeng, Saad, Yousef 2010. GPU-accelerated preconditioned iterative linear solvers. Minnesota Supercomputer Institute Report umsi-2010-112, University of Minnesota, Minneapolis, MN, 24 p.

\item Liang, Y., Weston, J., and Szularz, M. 2002. Generalized least-squares polynomial preconditioner for symmetric indefinite linear equations, Parallel Computing 28: 323-341.

\item McDonald, M.G., Harbaugh, A.W. 1988. A modular three-dimensional finite-difference ground-water flow model. U.S. Geological Survey Techniques of Water-Resources Investigations, book 6, chap. A1, 586 p.

\item Mehl, S.E., Hill, M.C. 2001. MODFLOW-2000, the U.S. Geological Survey modular ground-water model -- User guide to the LINK-AMG (LMG) Package for solving matrix equations using an algebraic multigrid solver. U.S. Geological Survey Open-File Report 01-177, 33 p.

\item Naff, R.L. 2008. Technique and Application of a Parallel Solver to MODFLOW, Proceedings of MODFLOW and More 2008: Ground Water and Public Policy, v. 5, May 19-21, 2008, Colorado School of Mines, Golden, Colorado.

\item Naff, R.L., Banta, E.R. 2008. The U.S. Geological Survey modular ground-water model—PCGN: A preconditioned conjugate gradient solver with improved nonlinear control. U.S. Geological Survey Open-File Report 2008-–1331, 35 p.

\item Niswonger, R.G., Panday, Sorab, and Ibaraki, Motomu 2011. MODFLOW-NWT, A Newton formulation for MODFLOW-2005. U.S. Geological Survey Techniques and Methods 6-A37, 44 p.

\item NVIDIA 2011. NVIDIA CUDA™: NVIDIA CUDA C Programming Guide Version 4.1.

\item OpenMP Architecture Review Board. 2005. OpenMP Application Program Interface, Version 2.5, accessed March 17, 2011 at \url{http://www.openmp.org/mp-documents/spec25.pdf}.

\item Petersen, W.P. and Arbenz, Peter 2004. Introduction to Parallel Computing: A Practical Guide with Examples in C. New York, Oxford University Press, 288 p.

\item Saad, Yousef 2003. Iterative methods for sparse linear systems – 2nd edition, Philadelphia, Pennsylvania, SIAM, 528 p.

\item Scandrett, C. 1989. Comparison of several iterative techniques in the solution of symmetric banded equations on a two-pipe Cyber 250, Applied Mathematics and Computation 34: 95-112.

\item Soni, Kunal, Chandar, D.D.J., and Sitaraman, J. 2012. Development of an overset grid computational fluid dynamics solver on graphical processing units. Computers \& Fluids 58: 1-4. doi: 10.1016/j.compfluid.2011.11.014.

\item Wilson, J.D., Naff, R.L. 2004. The U.S. Geological Survey modular ground-water model -- GMG linear equation solver package documentation. U.S. Geological Survey Open-File Report 2004--1261, 47 p.

\item Zheng, Chunmiao and Wang, P.P. 1999 MT3DMS. A modular three-dimensional multi-species transport model for simulation of advection, dispersion and chemical reactions of contaminants in groundwater systems; documentation and user's guide. U.S. Army Engineer Research and Development Center Contract Report SERDP-99-1, Vicksburg, MS, 202 p.

\end{description}

%GPGPU jargon  boxes
\newpage
%box 1
\fbox{
\parbox[c]{5.75in}{\label{GPGPUbox1}
\color{blue}\textbf{\\Box 1\\Developing Software for GPGPUs}

To use GPGPUs, specialized software must be developed to use a precompiled set of functions and routines which serve as a communication mechanism between a CPU and one or more GPGPU and pass instructions to the GPGPU.  These precompiled functions and routines are typically bundled into a code library that is accessed using a application programming interface (API).

\hspace{15pt}Kernels are blocks of code which are executed only on the GPGPU.  When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions. The kernel executed by the UPCG solver on the GPGPU is the preconditioned conjugate gradient algorithm.\\ \color{black}}}
\bigskip

\newpage
%box 2
\fbox{
\label{GPGPUbox2}
\parbox{5.75in}{
\color{blue}\textbf{\\Box 2\\NVIDIA GPGPU Architecture}

A GPGPU is a set of streaming multiprocessors (SM), each having many CUDA cores (CC); each CC is capable of executing many threads simultaneously. This nested architecture gives rises to the massively parallel capability of GPGPUs.

\hspace{15pt}A CUDA kernel function is executed on a CUDA grid, which is a grouping of threads. Each CUDA grid is composed of one or more blocks that are executed on each CC. For increased throughput, threads within a block are further split into warps, which is the maximum number of threads that are managed and processed simultaneously on each CC. Multiple warps execute one after another to complete block processing. All CCs within a SM share a single instruction, which results in massively parallel processing potential. GPGPU architecture is shown graphically in  \textbf{figure~\ref{FigParallel}}.

\hspace{15pt}The Tesla C2050 used in this study has the following specifications: SMs = 14, CC per SM = 32, threads per block = 1,024, and warp size = 32. The multi-threaded instruction unit on each of the 14 Tesla C2050 SMs can run 48 warps concurrently, yielding a total of 21,504 concurrent threads (14 SMs $\times$ 48 warps per SM $\times$ 32 threads per warp).\\ \color{black}
}}

%%Tables and Figures
%Table1
\newpage
\begin{table}[hp]
   \caption{Speedup of UPCG solver GPGPU simulations relative to CPU simulations for the same UPCG preconditioner. Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
   \label{TableGPUCPUSpeedup}
   \begin{tabular}{ l | c | c | c |}
 		\cline{2-4}
		& \multicolumn{3}{c|}{Speedup} \\ \cline{1-4}
		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & Jacobi & MILU0 & GLSPOLY \\ \cline{1-4}
        		\multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } &  1.513 &  0.990 &  2.972 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } &  9.536 &  1.699 & 16.315 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 18.871 &  1.404 & 26.786 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 24.453 &  1.938 & 33.615 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \textbf{26.918} &  1.936 & 35.154 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } &  3.234 &  1.517 &  6.146 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } & 13.420 &  1.742 & 18.665 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 19.877 &  1.322 & 25.420 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 23.871 &  1.861 & 29.516 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } &  5.234 &  1.517 &  9.231 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } & 15.848 &  1.769 & 20.843 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } & 22.444 &  1.265 & 27.658 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } & 25.171 &  1.870 & 30.760 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 11.552 &  1.715 & 17.564 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 20.806 &  1.833 & 25.826 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 24.981 &  1.889 & 29.888 \\ \cline{1-4}
       \end{tabular}
   \end{table}

%Table2
\newpage
\begin{table}[hp]
   \caption{Speedup of UPCG solver GPGPU simulations relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner. CPU times for the PCG solver with the MIC preconditioner are in seconds. Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
   \label{TableGPUSpeedup}
   \begin{tabular}{ l | c | c | c | c |}
 		\cline{2-5}
		& CPU Time & \multicolumn{3}{c|}{Speedup} \\ \cline{1-5}
		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & MIC & Jacobi & MILU0 & GLSPOLY \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } & 0.6826875 &  0.487 &  1.101 &  0.841 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } & 10.27194 &  1.675 &  1.543 &  2.294 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 75.64222 &  3.241 &  1.239 &  2.957 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 536.2608 &  1.891 &  1.438 &  2.754 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & 3891.683 & \textbf{ 1.357} &  1.696 &  2.839 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 1.683125 &  1.081 &  1.330 &  1.633 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } & 24.66269 &  3.156 &  1.534 &  3.497 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 175.7042 &  4.111 &  1.156 &  3.532 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 1058.137 &  1.967 &  1.590 &  2.837 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } & 3.040594 &  1.875 &  1.487 &  2.349 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } & 38.59184 &  3.908 &  1.530 &  4.075 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } & 315.6253 &  5.105 &  1.130 &  4.428 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } & 1962.365 &  2.560 &  1.634 &  3.562 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 18.50184 &  5.344 &  1.524 &  4.921 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 218.9169 &  8.683 &  1.598 &  7.643 \\ \cline{1-5}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 1621.158 &  9.299 &  1.621 &  8.633 \\ \cline{1-5}
       \end{tabular}
   \end{table}

%Table3
\newpage
\begin{table}[hp]
   \caption{Speedup of UPCG solver OpenMP simulations, with 4, 7, and 14 threads, relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner. Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
   \label{TableOpenMPSpeedup}
   \begin{tabular}{ l | c | c | c |}
 		\cline{2-4}
		& \multicolumn{3}{c|}{Speedup -- 4 / 7 /14 Threads} \\ \cline{1-4}
		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & Jacobi & MILU0 & GLSPOLY \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } & 1.166/0.614/0.614 &  2.446/1.511/1.423 & 1.185/0.779/1.241 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } & 0.447/0.332/0.357 & 1.637/1.197/1.164 & 0.571/0.364/0.535 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 0.409/0.419/0.359 & 1.488/1.463/1.194 & 0.313/0.323/0.309 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 0.185/0.182/0.151 & 1.691/1.622/1.286 & 0.225/0.272/0.198 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \textbf{0.111}/\textbf{0.099}/\textbf{0.108} & 1.619/1.523/1.670 &  0.222/0.241/0.245 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 1.113/1.090/1.004 & 1.954/2.258/1.567 & 1.171/1.147/1.626 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } &  0.536/0.567/0.460 & 1.500/1.521/1.151 & 0.525/0.625/0.556 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 0.527/0.472/0.411 & 1.543/1.474/1.227 & 0.396/0.295/0.335 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 0.211/0.198/0.168 & 1.401/1.693/1.418 & 0.237/0.313/0.244 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } & 0.872/1.067/0.897 & 1.433/1.901/1.300 & 0.827/1.080/1.346 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } &  0.577/0.482/0.527 & 1.444/1.417/1.328 & 0.548/0.579/0.598 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } &  0.593/0.562/0.479 & 1.519/1.590/1.377 & 0.389/0.340/0.385 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } &  0.245/0.246/0.215 & 1.746/1.478/1.544 & 0.313/0.286/0.295 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 1.066/1.059/1.215 & 1.503/1.574/1.426 & 0.858/0.899/1.186 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 0.970/1.014/1.019 & 1.563/1.604/1.400 & 0.780/0.799/0.913 \\ \cline{1-4}
        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 0.835/0.968/0.928 & 1.720/1.734/1.651 & 0.815/0.798/0.914 \\ \cline{1-4}
       \end{tabular}
   \end{table}

%Figure 1
\newpage
\begin{figure}[hp]
 	\centering
  	\includegraphics[width=17.15cm]{Figure1.png}
	\caption{(A) Fork-join parallel model. The master thread executes sequential operations and initiates thread forks used in parallel regions. (B) GPGPU architecture showing the relation of streaming multiprocessors and CUDA cores. The fork-join parallel model is applied to blocks of code being executed by thread processing units on multiple streaming multiprocessors.}
	\label{FigParallel}
\end{figure}

%Figure 2
\newpage
\begin{figure}[hp]
	\centering
  	\includegraphics[width=8.25cm]{Figure2.png}
 	\caption{Model domain showing the distribution of hydraulic conductivity and boundary conditions applied to a hypothetical unconfined aquifer. A total of 1,000 m$^{3}$/day are withdrawn from 4 pumping wells. Steady-state groundwater head contours (1 m contour interval) are also shown. Note the effect that groundwater withdrawals and hydraulic conductivity are having on simulated groundwater heads.}
	\label{FigModelDomain}
\end{figure}

%Figure 3
\newpage
\begin{figure}[hp]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure3a.png}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure3b.png}
	\end{subfigure}
 	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to sequential UPCG solver simulations executed on the CPU. Note the different scales used for MILU0 and GLSPOLY speedup.}
	\label{FigCPUGPUResults}
\end{figure}

%Figure 4
\newpage
\begin{figure}[hp]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure4a.png}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure4b.png}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure4c.png}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure4d.png}
	\end{subfigure}
 	\caption{Time, in seconds, spent applying the preconditioner, performing BLAS level 1 operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner. Note operation execution times have been normalized using the maximum preconditioner application time for each problem to facilitate comparison of different model runs.}
	\label{FigBLASOperationResults}
\end{figure}

%Figure 5
\newpage
\begin{figure}[hp]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure5a.png}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
  		\includegraphics[width=8.25cm]{Figure5b.png}
	\end{subfigure}
 	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}
	\label{FigGPUResults}
\end{figure}

%Figure 6
\newpage
\begin{figure}[hp]
	\centering
  	\includegraphics[width=8.25cm]{Figure6.png}
 	\caption{Speedup of parallel CPU simulations using the UPCG solver with the MILU0 preconditioner and 7 threads relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}
	\label{FigOpenMPResults}
\end{figure}

\end{document}

### 2012-08-27 ResponseTo_GW20120518-0099_InternalReview.tex
Latest revisions to document

\documentclass[12pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

\usepackage{geometry}
\geometry{letterpaper}
\geometry{margin=1in}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{color}

\usepackage{amsmath}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{hyperref}
\renewcommand{\UrlBreaks}{\do\/\do\a\do\b\do\c\do\d\do\e\do\f\do\g\do\h\do\i\do\j\do\k\do\l\do\m\do\n\do\o\do\p\do\q\do\r\do\s\do\t\do\u\do\v\do\w\do\x\do\y\do\z\do\A\do\B\do\C\do\D\do\E\do\F\do\G\do\H\do\I\do\J\do\K\do\L\do\M\do\N\do\O\do\P\do\Q\do\R\do\S\do\T\do\U\do\V\do\W\do\X\do\Y\do\Z}

\begin{document}

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
Internal reviews were performed by (1) Dr. Yong Liu a senior research scientist at the National Center for Supercomputing Applications (NCSA) University of Illinois at Urbana-Champaign, (2) Randy Hanson a research hydrologist in the California Water Science Center, and (3) Vevek Bedekar a professional engineer with  S.S. Papadopulos \& Associates, Inc. Randy Hanson provided a few comments in a review memorandum, which are included and addressed below, and as inline comments in the manuscript pdf; inline comments are addressed in the document titled MODFLOW-GPU\_03.jdhResponse.pdf. Internal review comments were helpful and have been addressed in the revised manuscript. Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.

\section*{\small{Yong Liu Comments:}}
\begin{enumerate}
\item \textit{One thing that I would suggest to add is to say a few words about any possible future research (e.g., GPU cluster, Cloud-based GPU resources such as Amazon GPU cluster) and practical usage of GPU solvers for MODFLOW 2005 for real-world use cases (an example real-world use case that would benefit from using this research findings and the GPU solver would be excellent!.}

Response

\item \textit{It would be also nice if the authors can indicate where and how to obtain the source code (will it be in public domain just like the MODFLOW source code?.}

Response

\item \textit{It would be nice if you could also provide the python code so that I or someone else can reproduce some of your experiments.}

Response

\end{enumerate}

\section*{\small{Randy Hanson Comments:}}
\begin{enumerate}
\item \textit{In particular the preconditioning seems to be an issue and one is left wondering if the code could be written better or is it simply a hardware issue given the structure of the code. I have made several inline comments that may give more details as to where these issues might require minor additional elaboration or insight based on your experience. Would a graph showing the change in performance while varying the inner and outer iterations for one of the models provide any additional insight on how to best set the combination of these two, or is this too problem dependent?}

Response.

\item \textit{The reader should also have a clear idea of how he could implement these choices and which choice is best for the problem at hand to some degree (although this may be outside the scope of this paper).}

Response.

\item \textit{The reader may also wonder if additional compilation features such as provided by Intel compiler for memory alignment, vectorization, use of coarrays, use of math-kernal library vectorized functions, or chip specific architectures could provide additional performance enhancements.}

Response.

\item \textit{Inline comments in pdf  MODFLOW-GPU\_03.pdf.}

See responses in MODFLOW-GPU\_03.jdhResponse.pdf.

\end{enumerate}

\section*{Vevek Bedekar Comments:}
\begin{enumerate}
\item \textit{Specify whether this work is first of its kind in applying GPUs in the field of groundwater modeling or specifically in the application of MODFLOW. Otherwise, cite other works that apply GPUs to solve groundwater simulations. The following two search results were found as a result of a quick Google search.}

\textit{\href{http://doi.acm.org/10.1145/1774088.1774588}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, A simulation of large-scale groundwater flow on CUDA-enabled GPUs: Proceedings of the 2010 ACM Symposium on Applied Computing, 2402--2403. doi: 10.1145/1774088.1774588}}

\textit{\href{http://www.gpucomputing.net/?q=node/14182}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, CUDA-based solver for large-scale groundwater flow simulation: Engineering with Computers, 28(1), 13--19. doi: 10.1007/s00366-011-0213-2}}

The second reference (Ji \textit{et al.}, 2012) is a reference in the manuscript but this is difficult to determine from the link provided. The first reference (Ji \textit{et al.}, 2012) was not included in the original manuscript since it is a conference proceeding and may be difficult for some readers to locate. It has been added to ensure Ji \textit{et al.}, 2010 is credited as being the first application of solving MODFLOW using GPGPUs.

\end{enumerate}

\end{document}

### 2012-08-27 ResponseTo_GW20120518-0099_Review.tex
Latest revisions to document

@@
 
-Response.
+Additional discussion of ideal parallelization and issues that make achieving ideal speedup has been added to the introduction.
 
@@
 
-Response.
+We have attempted to eliminate all but the essential GPGPU jargon in the main text.  A significant portion of the jargon has been moved to Boxes 1 and 2.
 
@@
 
-A section titled ``How to Obtain the Software'' has been added that provides information on how to obtain the software.
+A section titled ``\textbf{How to Obtain the Software}'' has been added that provides information on how to obtain the software.
+The source code and a precompiled Windows executable will be provided.
+Instructions for compiling the source code have been provided in the ``\textbf{How to Obtain the Software}''  section.
 
@@
 \item \textit{GPGPU speedup has been compared to the PCG5 solver on a serial CPU.
-With more advanced solvers coming into the MODFLOW family (for example the XMD solver in MODFLOW-NWT,)it would be interesting to note the speedup of UPCG compared to even more robust higher level ILU solvers which may be even faster than PCG5 for hard problems.}
+With more advanced solvers coming into the MODFLOW family (for example the XMD solver in MODFLOW-NWT), it would be interesting to note the speedup of UPCG compared to even more robust higher level ILU solvers which may be even faster than PCG5 for hard problems.}
 
-The analyses presented in this manuscript are not meant to be an exhaustive comparison of GPGPU solutions but an evaluation of the speedup possible.
+The authors agree with the reviewer that higher level ILU preconditioners may be more effective than the standard MODFLOW PCG solver for particular types of problems.
+The analyses presented in this manuscript are not meant to be an exhaustive comparison of GPGPU solutions but an evaluation of the speedup possible when compared to a standard MODFLOW solver that uses a similar solution approach.
 
@@
 
-Response.
+The authors disagree with this comment.
+We believe the issues related to solution of the groundwater flow equation on a GPGPU are presented in the relevant locations and important issues are summarize again in the ``\textbf{Conclusions}'' section.
 
@@
 
-Response.
+We have tried to eliminate all non-essential GPGPU jargon and make the main text more readable for the typical Ground Water reader.
+A significant portion of the jargon has been moved to Boxes 1 and 2 as suggested by the editor.
 
@@
 
-Response.
+Suggested modification made.
 
@@
 
-Response.
+The font sizes have been increased on all of the figures.
+The symbol size and lines on \textbf{figures 3}, \textbf{5}, and \textbf{6} have been increased and converted to solid colored lines, respectively, to improve the readability as suggested.
 
@@
 
-Response.
+Suggested addition made.
 
@@
 
-Response.
+Yes.
 
@@
 
-Response.
+CUBLAS and BLAS have been defined.
 

### 2012-08-27 GW20120518-0099-r01.tex
Latest revisions to document

--- 2012-08-02 GW20120518-0099.tex
@@
 In addition to OpenMP and MPI, GPGPUs also presents an option for linear solver parallelization.
-In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs).
+In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 \color{blue} cores \color{black} in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs).
 Use of GPGPUs to parallelize linear solvers for MODFLOW-2005 is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
@@
 Although the connectivity of the finite-difference discretization used in MODFLOW-2005 uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
-This storage format is also supported by the CUBLAS libraries.
+This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS).
+BLAS is a library of Basic Linear Algebra Subprograms\color{black}.
 The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Sadd 2003).
@@
 Head-dependent (general head) boundaries were specified on the left and right sides and no flow boundaries at the top and bottom.
-Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain $\left [ \left(10.
-- 0.0\right) / 1000.
-= 0.01 \text{ m/m} \right]$ and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively.
+Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10.
+/ 1000.
+= 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively.
 For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively.
@@
 
-The PCG algorithm was implemented for use on GPGPUs using the NVIDIA CUDA-enabled implementation of the BLAS library (CUBLAS).
+The PCG algorithm was implemented for use on GPGPUs using the NVIDIA \color{blue}CUBLAS library\color{black}.
 The availability of this API makes it relatively simple to develop parallel GPGPU code.
@@
 
-The UPCG solver was developed to solve the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners.
+The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners.
 An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library.

### 2012-08-30 ResponseTo_GW20120518-0099_Review.tex
Incremental modifications

@@
 
-Response.
+Suggested modification made. Also MODFLOW-2005 has been replaced with MODFLOW in the revised manuscript except where it is necessary to specifically discuss the current version of MODFLOW.
 
@@
 
-Response.
+Correction made.
 
@@
 
-Response.
+A significant portion of the jargon has been moved to Boxes 1 and 2.
 
@@
 
-Response.
+We have added text to indicate that GPGPU parallelization has great potential for MODFLOW and other numerical simulators.
 
@@
 
-Response.
+Figure 4 has been modified as suggested.
 

### 2012-08-30 GW20120518-0099-r01.tex
Incremental modifications

@@
 \renewcommand{\headrulewidth}{0pt}
-\lhead{}\chead{Hughes and White -- MODFLOW-2005 UPCG GPGPU Solver}\rhead{}
+\lhead{}\chead{Hughes and White -- MODFLOW UPCG GPGPU Solver}\rhead{}
 \lfoot{}\cfoot{\thepage}\rfoot{}
 
-\title{Use of general purpose graphical processing units with MODFLOW-2005}
+\title{Use of general purpose graphical processing units with MODFLOW}
 \author{Hughes, J.D.\\U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jdhughes@usgs.gov} \and White, J.T. \\ U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jwhite@usgs.gov}}
@@
 \begin{abstract}
-To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW-2005, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed.
+To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed.
 The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete LU factorization, and generalized least-squares polynomial preconditioners.
@@
 The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells.
-Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW-2005 PCG solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
-Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be parallelized.
+Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW PCG solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
+Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
 \end{abstract}
@@
 
-MODFLOW-2005 (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to two- and three-dimensional problems since 1984.
-Because MODFLOW-2005 uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
+MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to \color{blue}numerous \color{black}two- and three-dimensional problems.
+Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
 To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
@@
 Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7.
-In addition to OpenMP and MPI, GPGPUs also presents an option for linear solver parallelization.
+In addition to OpenMP and MPI, GPGPUs also \color{blue}present \color{black}an option for linear solver parallelization.
 In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 \color{blue} cores \color{black} in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs).
-Use of GPGPUs to parallelize linear solvers for MODFLOW-2005 is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
+Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
 For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems.
@@
 The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.
-\color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} not just for MODFLOW-2005, but for most numerical modeling applications\color{black}.
+\color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} not just for MODFLOW, but for most numerical modeling applications\color{black}.
 
@@
 
-MODFLOW-2005 uses a cell-centered, finite-difference approximation of \textbf{equation~\ref{gwpde}} to solve for three-dimensional groundwater flow (Harbaugh 2005).
+MODFLOW uses a cell-centered, finite-difference approximation of \textbf{equation~\ref{gwpde}} to solve for three-dimensional groundwater flow (Harbaugh 2005).
 Development of the finite-difference form of \textbf{equation~\ref{gwpde}} using the continuity equation for a cell results in
@@
 
-Although the connectivity of the finite-difference discretization used in MODFLOW-2005 uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
+Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
 This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS).
@@
 Infinity norms ($\left \|  x \right \|_{\infty}$) are used to evaluate convergence of the UPCG linear solver with respect to the change in $\mathbf{h}$ (HCLOSE) and $\mathbf{r}$ (RCLOSE).
-In MODFLOW-2005, non-linearities are resolved using Picard iteration.
+In MODFLOW, non-linearities are resolved using Picard iteration.
 Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver.
@@
 \section* {\bf Discussion}
-Although notable, CPU/GPGPU speedups summarized on \textbf{table~\ref{TableGPUCPUSpeedup}} are not a fair measure of the speedups that would result from use of the GPGPU to solve a MODFLOW-2005 problem.
-A more realistic comparison is the speedup of GPGPU simulations relative to sequential CPU simulations using the standard MODFLOW-2005 PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
+Although notable, CPU/GPGPU speedups summarized on \textbf{table~\ref{TableGPUCPUSpeedup}} are not a fair measure of the speedups that would result from use of the GPGPU to solve a MODFLOW problem.
+A more realistic comparison is the speedup of GPGPU simulations relative to sequential CPU simulations using the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
 PCG(MIC)/GPGPU speedups for the Jacobi, MILU0, and GLSPOLY preconditioners are summarized on \textbf{table~\ref{TableGPUSpeedup}}.
@@
 PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners range from less than 1.0 to more than 8.0, which are notably less than CPU/GPGPU speedups.
-This is attributable to the efficiency of the standard MODFLOW-2005 PCG solver with the MIC preconditioner.
+This is attributable to the efficiency of the standard MODFLOW PCG solver with the MIC preconditioner.
 The relative increase in PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).
@@
 PCG(MIC)/UPCG-OpenMP speedups for the MILU0 preconditioner with 7 OpenMP threads are shown graphically on \textbf{figure~\ref{FigOpenMPResults}}.
-It should be noted that the Intel Fortran compiler (version 12.1.2.278), with the highest-level optimization option (/O3 compiler switch), was used to compile MODFLOW-2005 with the UPCG solver.
+It should be noted that the Intel Fortran compiler (version 12.1.2.278), with the highest-level optimization option (/O3 compiler switch), was used to compile MODFLOW with the UPCG solver.
 The relatively small PCG(MIC)/UPCG-OpenMP speedups observed with the UPCG solver compiled using the Intel Fortran compiler is consistent with the observations of Dong and Li (2009).
@@
 Unlike fork-join approaches used on CPUs (OpenMP), which have thread limits determined by the number of CPUs and cores per CPU, GPGPUs have the capability of using thousands to tens of thousands of threads concurrently in parallel sections.
-The number of concurrent threads that can be used on GPGPUs makes them ideal for parallelization of simultaneous solutions of systems of linear equations like those solved by MODFLOW-2005.
+The number of concurrent threads that can be used on GPGPUs makes them ideal for parallelization of simultaneous solutions of systems of linear equations like those solved by MODFLOW.
 The key to successful parallelization on GPGPUs is to (1) use numerical approaches that are highly parallelizable and (2) reduce the need for memory transfers between the CPU and GPGPU as much as possible.
@@
 
-A number of test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW-2005.
+A number of test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW.
 CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared to CPU runtimes for the same preconditioner.
@@
 
-Speedups calculated using GPGPU runtimes and CPU runtimes for the same solver are not fair measures of the benefits of using GPGPUs to solve the groundwater flow equation because of the efficiency of the standard MODFLOW-2005 PCG solver with the MIC preconditioner.
+Speedups calculated using GPGPU runtimes and CPU runtimes for the same solver are not fair measures of the benefits of using GPGPUs to solve the groundwater flow equation because of the efficiency of the standard MODFLOW PCG solver with the MIC preconditioner.
 PCG(MIC)/GPGPU speedups are smaller than speedups calculated using sequential CPU runtimes for the UPCG solver.
-Use of GPGPUs with the UPCG solver and the GLSPOLY preconditioner can result in PCG(MIC)/GPGPU speedups as large as 8 for large multi-layer models when compared to the standard MODFLOW-2005 PCG solver with the MIC preconditioner.
+Use of GPGPUs with the UPCG solver and the GLSPOLY preconditioner can result in PCG(MIC)/GPGPU speedups as large as 8 for large multi-layer models when compared to the standard MODFLOW PCG solver with the MIC preconditioner.
 For certain problems, the Jacobi preconditioner is adequate and has similar PCG(MIC)/GPGPU speedups.
@@
 
-Source code for MODFLOW--2005 with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
+Source code for MODFLOW (MODFLOW--2005 version) with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
 Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit which includes the necessary CUDA drivers.
@@
 It is possible that the software may need to be modified for application of the code on GPGPU models other than the Tesla C2050.
-And finally, python scripts used to create the MODFLOW-2005 datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
+And finally, python scripts used to create the MODFLOW datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
 

### 2012-08-31 ResponseTo_GW20120518-0099_Review.tex
Additional edits

@@
 
-Response.
+This sentence has been revised to indicate that GPGPUs are capable of executing more floating point operations per second than CPUs.
 
@@
 
-Response.
+This sentence has been rephrased to clarify the description of the thickness of layer 1.
 
@@
 
-\item \textit{Line376 - as indicated above, can you provide some guidance here about when one approach might be better than another depending on the system properties (size, K field).
+\item \textit{Line 376 - as indicated above, can you provide some guidance here about when one approach might be better than another depending on the system properties (size, K field).
 Also, GMG should be included here to so that the full suite of approaches are compared (weak but highly parallelizable to strong but not very parallelizable).
@@
 
-\item \textit{As indicated above lines 75-108 are filled with jargon and should be reconsidered. 
-Response.}
-
-Response.
-
 \end{enumerate}
@@
 
-Response.
+Double-precision floating point variables are used in the solvers and the tables have been rounded to single-precision numbers.
+Since we are reporting results from a numerical analysis we thought it appropriate to retain a larger number of digits to allow readers the ability to evaluate our results differently than we have.
+For example it would be possible for readers to calculate accurate run times (in seconds) for each of the simulations summarized in \textbf{table 2} using the speedup values and MIC cpu times.
+This may be of benefit for future studies conducted by other researchers.
+However, we would be willing to modify the number of digits reported if the editor agrees with this comment.
 
@@
 
-Response.
+\textbf{Table 4} has been added which includes information on the number of iterations required for each of the solvers and various preconditioners.
+Additional discussion has been added in the \textbf{Discussion} section to provide some guidance on the strength of the preconditioners for the problems evaluated.
 
@@
 
-Response.
+This is primarily based on the increase in the number of simultaneous linear equations being solved.
+Differences in the \textbf{A} matrix resulting from different cell sizes can also affect solver performance.
+Additional clarification has been added.
 
@@
 
-Response.
+Additional discussion of the reasons the particular test cases were evaluated has been provided.
 

### 2012-08-31 GW20120518-0099-r01.tex
Additional edits

@@
 \usepackage{lineno}
-\usepackage{color}
+\usepackage{color, colortbl}
 \usepackage[kernelfbox]{xcolor}
@@
 In addition to OpenMP and MPI, GPGPUs also \color{blue}present \color{black}an option for linear solver parallelization.
-In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 \color{blue} cores \color{black} in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs).
-Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
+In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and as a result are capable of executing more floating point operations per second.
+\color{black}Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
 For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems.
@@
 The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.
-\color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} not just for MODFLOW, but for most numerical modeling applications\color{black}.
+\color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations\color{black}.
 
@@
 For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers.
-The thickness of layer 1 was calculated to be sum of the top elevation of the model (10 m) and constant thickness used for layers 2 and below.
+\color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.
 
-The heterogeneous hydraulic conductivity distribution (base hydraulic conductivity data) was generated for the 1,000 $\times$ 1,000 cell model (base model) using sequential Gaussian simulation with log-transformed (base 10) parameters based on a mean value of 5 m/d and an exponential variogram having a nugget of 1.023 $\log_{10}((m/d)^{2})$, a variance of 1.23 $\log_{10}((m/d)^{2})$, and length parameter ($a$) of 250.5 m (effective range $\approx$ 750 m).
+\color{black}The heterogeneous hydraulic conductivity distribution (base hydraulic conductivity data) was generated for the 1,000 $\times$ 1,000 cell model (base model) using sequential Gaussian simulation with log-transformed (base 10) parameters based on a mean value of 5 m/d and an exponential variogram having a nugget of 1.023 $\log_{10}((m/d)^{2})$, a variance of 1.23 $\log_{10}((m/d)^{2})$, and length parameter ($a$) of 250.5 m (effective range $\approx$ 750 m).
 The axis of anisotropy was rotated by 45 degrees to the model grid and a 3.5:1 ratio of horizontal anisotropy was used.
@@
    \caption{Speedup of UPCG solver GPGPU simulations relative to CPU simulations for the same UPCG preconditioner.
-Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
+Speedup values in shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
    \label{TableGPUCPUSpeedup}
@@
         \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 24.453 &  1.938 & 33.615 \\ \cline{1-4}
-        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \textbf{26.918} &  1.936 & 35.154 \\ \cline{1-4}
+        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \cellcolor{gray} 26.918 &  1.936 & 35.154 \\ \cline{1-4}
         \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } &  3.234 &  1.517 &  6.146 \\ \cline{1-4}
@@
 CPU times for the PCG solver with the MIC preconditioner are in seconds.
-Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
+Speedup values in  in shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
    \label{TableGPUSpeedup}
@@
         \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 536.2608 &  1.891 &  1.438 &  2.754 \\ \cline{1-5}
-        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & 3891.683 & \textbf{ 1.357} &  1.696 &  2.839 \\ \cline{1-5}
+        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & 3891.683 & \cellcolor{gray} 1.357 &  1.696 &  2.839 \\ \cline{1-5}
         \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 1.683125 &  1.081 &  1.330 &  1.633 \\ \cline{1-5}
@@
    \caption{Speedup of UPCG solver OpenMP simulations, with 4, 7, and 14 threads, relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.
-Speedup values in \textbf{bold} type indicate simulations that did not converge within 50 outer iterations with up to 1,000 inner iterations.}
+Speedup values in in shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
    \label{TableOpenMPSpeedup}
@@
         \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 0.185/0.182/0.151 & 1.691/1.622/1.286 & 0.225/0.272/0.198 \\ \cline{1-4}
-        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \textbf{0.111}/\textbf{0.099}/\textbf{0.108} & 1.619/1.523/1.670 &  0.222/0.241/0.245 \\ \cline{1-4}
+        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \cellcolor{gray}0.111/0.099/0.108 & 1.619/1.523/1.670 &  0.222/0.241/0.245 \\ \cline{1-4}
         \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 1.113/1.090/1.004 & 1.954/2.258/1.567 & 1.171/1.147/1.626 \\ \cline{1-4}
@@
 
+%Table4
+
+\newpage
+\begin{table}[hp]
+\caption{Convergence summary for simulations using the PCG solver with the MIC preconditioner, the GMG solver with the ILU0 preconditioner, and the UPCG solver with the Jacobi, MILU0, and GLSPOLY preconditioners.
+The total number of outer and inner iterations are shown (outer/inner).
+Shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
+   \label{tab:UPCGITSummary}
+   \begin{tabular}{ l | c | c | c | c | c |}
+ 		\cline{2-6}
+		& \multicolumn{5}{c|}{Convergence History} \\ \cline{1-6}
+		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & MIC & GMG & Jacobi & MILU0 & GLSPOLY \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } & 6/140 & 5/7 & 5/955 & 6/136 & 6/130 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } & 7/312 & 5/9 & 6/2629 & 7/298 & 7/352 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 9/574 & 6/10 & 8/4624 & 8/547 & 7/765 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 8/1022 & 7/14 & 21/17862 & 8/980 & 8/1752 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & 9/1859 & 8/16 & \cellcolor{gray}50/50000 & 9/1758 & 9/3257 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 6/151 & 5/7 & 5/811 & 6/145 & 5/123 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } & 6/334 & 7/12 & 5/2073 & 6/315 & 5/276 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 7/597 & 8/16 & 6/4102 & 7/562 & 6/633 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 7/902 & 20/33 & 17/15342 & 7/864 & 5/1363 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } & 6/167 & 5/6 & 5/788 & 5/154 & 5/124 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } & 5/335 & 6/11 & 5/1967 & 5/319 & 5/261 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } & 7/685 & 7/14 & 6/4059 & 6/633 & 5/637 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } & 6/1069 & 23/33 & 16/14848 & 6/1003 & 5/1359 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 6/291 & 5/7 & 5/955 & 6/273 & 5/164 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 5/546 & 6/9 & 5/1908 & 5/512 & 5/275 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 7/1014 & 7/13 & 6/3913 & 6/955 & 5/521 \\ \cline{1-6}
+       \end{tabular}
+   \end{table}
+
 %Figure 1
 \newpage
+
 \begin{figure}[hp]

### 2012-09-04 GW20120518-0099-r01.tex
Edits to manuscript and Randy Hanson's review documents

@@
 Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
-To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
-The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (SOR) linear solver.
-These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG2) linear solver (Hill 1990), the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
+\color{blue}Increase model runtimes are typically a direct result of increases in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
+\color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
+The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver.
+These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG2) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
 
@@
 However, ideal parallelization is rarely, if ever, achieved because of communication bottlenecks which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
+\color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}
 
@@
 For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems.
-For solution of the groundwater flow equation, Ji \textit{et al.} (2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.
+For solution of the groundwater flow equation, Ji \textit{et al.} (\color{cyan}2010 and \color{black}2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.
 
@@
 
-In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the residual of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless].
+In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the \color{cyan}difference between the left and right side (residual) \color{black}of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless].
+\color{cyan}Readers are referred to Barrett \textit{et al.} (1994) for more information on the PCG method.\color{black}
 
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize (Liang \textit{et al.} 2002; Saad 2003).
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number matrix-vector product and a constant times a vector plus a vector (axpy) BLAS operations \color{black} (Liang \textit{et al.} 2002; Saad 2003).
 
@@
 Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
-This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS).
-BLAS is a library of Basic Linear Algebra Subprograms\color{black}.
-The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Sadd 2003).
+This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012).
+BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979)\color{black}.
+The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (\color{cyan}Saad \color{black}2003).
 
@@
 For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and level-1 BLAS operations.
-The OpenMP matrix-vector product was constructed using a level-2 BLAS GEMV approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
+The OpenMP matrix-vector product was constructed using a level-2 BLAS \color{cyan}general matrix-vector (GEMV) product \color{black}approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
 OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively.
@@
 Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver.
-This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of outer (Picard) iterations is exceeded.
-\\
+This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of Picard (outer) iterations is exceeded; \color{cyan}the UPCG solver approach for evaluating convergence is identical to the standard MODFLOW PCG solver (Hill 1990).\color{black}
 
 \section* {\bf Test Cases}
-The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing perfomance (NVIDIA compute capability 2.0).
-The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel Xenon 3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS.
-For comparison with GPGPU results, parallel CPU simulations were executed on dual 4-core Intel Xenon 2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).
+The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing performance (NVIDIA compute capability 2.0).
+The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel \color{cyan}Xeon \color{black}3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS.
+For comparison with GPGPU results, \color{cyan}serial and \color{black}parallel CPU simulations were executed on dual 4-core Intel \color{cyan}Xeon \color{black}2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).
+
+\color{blue}To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes if discretized using different horizontal and vertical cell sizes.
+The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear as a result of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$.
+Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.\color{black}
 
@@
 For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers.
-\color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.
+\color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.\color{black}
 
-\color{black}The heterogeneous hydraulic conductivity distribution (base hydraulic conductivity data) was generated for the 1,000 $\times$ 1,000 cell model (base model) using sequential Gaussian simulation with log-transformed (base 10) parameters based on a mean value of 5 m/d and an exponential variogram having a nugget of 1.023 $\log_{10}((m/d)^{2})$, a variance of 1.23 $\log_{10}((m/d)^{2})$, and length parameter ($a$) of 250.5 m (effective range $\approx$ 750 m).
+The heterogeneous hydraulic conductivity distribution (base hydraulic conductivity data) was generated for the 1,000 $\times$ 1,000 cell model (base model) using sequential Gaussian simulation with log-transformed (base 10) parameters based on a mean value of 5 m/d and an exponential variogram having a nugget of 1.023 $\log_{10}((m/d)^{2})$, a variance of 1.23 $\log_{10}((m/d)^{2})$, and length parameter ($a$) of 250.5 m (effective range $\approx$ 750 m).
 The axis of anisotropy was rotated by 45 degrees to the model grid and a 3.5:1 ratio of horizontal anisotropy was used.
@@
 
-Head-dependent (general head) boundaries were specified on the left and right sides and no flow boundaries at the top and bottom.
+Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no flow boundaries at the top and bottom.
 Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10.
@@
 The head-dependent boundaries cause an ambient groundwater flow from left to right.
-Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d.
+Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d \color{cyan}and represents 46\% of the flow entering the model\color{black}.
 For multi-layer simulations, the pumping wells withdraw water from the lower model layer.
@@
 For all model discretizations, the UPCG solver with the Jacobi, MILU0, and GLSPOLY preconditioners were evaluated on the CPU, using both the sequential and parallel (OpenMP) options, and GPGPU.
-To maximize the amount of time spent in the linear solvers the maximum number of outer (Picard) and inner (linear) iterations were set to 50 and 1,000, respectively.
+To maximize the amount of time spent in the linear solvers the maximum number of Picard and inner (linear) iterations were set to 50 and 1,000, respectively.
 A HCLOSE value of 0.001 m was used for all simulations.
@@
 Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads.
-Simulated heads for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
+Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
+\color{cyan}Simulated water-levels for simulations that satisfy specified convergence criteria and use different (1) solvers, (2) preconditioners, and/or (3) hardware are within $\pm$RCLOSE m.\color{black}
 
 \subsection*{Results}
+\color{blue}The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and the combination of the GPGPU and CPU were identical to the number required on the CPU.
+The number of Picard and inner iterations required by the MILU0 preconditioner is comparable to the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
+The number of Picard iterations is also comparable between the MILU0, PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner; the GMG inner iterations are not comparable because of the differences in the inner iteration approaches of both the UPCG and PCG solvers and the GMG solver.
+The number of Picard and inner iterations required by the GLSPOLY preconditioner generally exceeds the MILU0 preconditioner and the PCG solver with the MIC preconditioner by a factor of but it is notably better the Jacobi preconditioner.
+In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.\color{black}
+
 The CPU/GPGPU speedup of UPCG solver simulations for the Jacobi, MILU0, and GLSPOLY preconditioners are summarized on \textbf{table~\ref{TableGPUCPUSpeedup}}.
@@
 Only marginal speedups, ranging from less than 1.0 to almost 2.0, are observed with the MILU0 preconditioner.
-The maximum number of outer (Picard) and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.
 The smallest CPU/GPGPU speedup corresponds with the smallest model discretizations (200 $\times$ 200 $\times$ 1) because of the overhead associated with GPGPU-CPU memory copy operations relative to the solution time.
@@
 For the MILU0 preconditioner, all of the speedup occurs during the conjugate gradient routine in the level-1 BLAS operations [matrix-vector multiplication and vector operations (dot products)], which are carried out on the GPGPU after application of the preconditioner.
-Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases.
+Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases \color{cyan}(\textbf{fig.~\ref{FigBLASOperationResults}a} and \textbf{b})\color{black}.
 For the GLSPOLY preconditioner, solution on the GPGPU reduces the time spent completing linear solver operations by more than 95\% and the time spent transferring data between the CPU and GPGPU is negligible.
@@
 Although notable, CPU/GPGPU speedups summarized on \textbf{table~\ref{TableGPUCPUSpeedup}} are not a fair measure of the speedups that would result from use of the GPGPU to solve a MODFLOW problem.
-A more realistic comparison is the speedup of GPGPU simulations relative to sequential CPU simulations using the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
+A more realistic comparison is the speedup of GPGPU simulations relative to sequential CPU simulations using the standard MODFLOW PCG solver with the MIC preconditioner (Hill 1990).
 PCG(MIC)/GPGPU speedups for the Jacobi, MILU0, and GLSPOLY preconditioners are summarized on \textbf{table~\ref{TableGPUSpeedup}}.
@@
 PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners range from less than 1.0 to more than 8.0, which are notably less than CPU/GPGPU speedups.
-This is attributable to the efficiency of the standard MODFLOW PCG solver with the MIC preconditioner.
-The relative increase in PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).
+\color{cyan}Reduced PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners relative to CPU/GPGPU speedups can be attributed to the general efficiency of the MODFLOW PCG solver that is a result of serial coding to reduce solver memory requirements.
+\color{black}The relative increase in PCG(MIC)/GPGPU speedups \color{cyan}observed for larger problems with \color{black}the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).
 
 Parallel CPU simulations were run to compare speedups possible with GPGPUs to speedups that could be realized on multi-core CPUs using the OpenMP programming paradigm.
-PCG(MIC)/UPCG-OpenMP speedups for the Jacobi, MILU0, and GLSPOLY preconditioners using 4, 7, and 14 OpenMP threads are summarized on \textbf{table~\ref{TableOpenMPSpeedup}}.
+\color{blue} The PCG solver parallelized by Dong and Li (2009) using OpenMP was not evaluated in this study because we wanted to directly compare CPU, GPGPU, and parallel CPU runtimes for the same solver.
+\color{black}PCG(MIC)/UPCG-OpenMP speedups for the Jacobi, MILU0, and GLSPOLY preconditioners using 4, 7, and 14 OpenMP threads are summarized on \textbf{table~\ref{TableOpenMPSpeedup}}.
 PCG(MIC)/UPCG-OpenMP speedups for the Jacobi and GLSPOLY preconditioners are notably less than PCG(MIC)/GPGPU speedups and is expected because the number of threads on the GPGPU is orders of magnitude greater than for parallel CPU simulations.
-For the MILU0 preconditioner, PCG(MIC)/UPCG-OpenMP speedups (average speedup $=$ 1.539) are comparable to PCG(MIC)/GPGPU speedups (average speedup $=$ 1.447) and there is little benefit to using the GPGPU with the MILU0 preconditioner.
+\color{cyan} In some cases, PCG(MIC)/UPCG-OpenMP speedups are less than 1 as a result of the limited number of threads available and the numerical overhead associated with managing OpenMP threads.
+\color{black}For the MILU0 preconditioner, PCG(MIC)/UPCG-OpenMP speedups (average speedup $=$ 1.539) are comparable to PCG(MIC)/GPGPU speedups (average speedup $=$ 1.447) and there is little benefit to using the GPGPU with the MILU0 preconditioner.
 PCG(MIC)/UPCG-OpenMP speedups for the MILU0 preconditioner with 7 OpenMP threads are shown graphically on \textbf{figure~\ref{FigOpenMPResults}}.
@@
 The relatively small PCG(MIC)/UPCG-OpenMP speedups observed with the UPCG solver compiled using the Intel Fortran compiler is consistent with the observations of Dong and Li (2009).
-Larger PCG(MIC)/UPCG-OpenMP speedups would likely be observed with other compilers but the value of using a less optimized compiler is questionable.
+\color{blue} We observed similar PCG(MIC)/UPCG-OpenMP speedups with and without optimization of the executable compiled by the Intel Fortran compiler.\color{black}
+
+\color{blue}Based on our results we believe we can offer the following suggestion regarding use of parallel strategies on GPGPUs and CPUs: (1) in cases where the GLSPOLY preconditioner is adequate, solution on the GPGPU will result in significant speedup over serial and parallel CPU solutions; (2) in some cases the use of the Jacobi preconditioner will result in a GPGPU speedup that exceeds the speedup possible with other preconditioner options but this will be the exception rather than the rule and only applicable to select problems (and likely very simple problems); (3) in cases where a stronger preconditioner than the GLSPOLY preconditioner is required, the parallel CPU version of the MILU0 preconditioner will result in the largest speedup; and (4) in cases with smooth, low-frequency error components a completely different solver, such as the LMG or GMG solver, may be needed (see Mehl and Hill 2001).
+The effectiveness of a preconditioner will be based on the distribution of the eigenvalues of the preconditioned matrix.
+For some problems the eigenvalues of the Jacobi and/or GLSPOLY preconditioned matrices will be sufficiently clustered to allow a significant reduction in the number of iterations required to achieve convergence.
+For other problems, a stronger incomplete factorization preconditioner will be needed to cluster the eigenvalues of the preconditioner matrix.
+In general for a MODFLOW application, it becomes more difficult for a preconditioner to cluster eigenvalues as the smoothness of the hydraulic conductivity field decreases.
+So it would be expected that the effectiveness of Jacobi and GLSPOLY preconditioners would decrease as the variance of the hydraulic conductivity field increases.
+For the test cases we evaluated, we found that the Jacobi and GLSPOLY preconditioners were adequate for hydraulic conductivity fields containing values spanning 3 orders of magnitude.
+The effectiveness of the Jacobi preconditioner was diminished (that is, required additional iterations) for hydraulic conductivity fields containing values spanning 5 orders of magnitude (results not shown).
+We were able to satisfy convergence criteria for all of the alternative problems evaluated because of the large number of iterations that were allowed (as many as 50,000 inner iterations).\color{black}
 
@@
 When compared to parallel CPU solutions, speedups for GPGPU solutions using the UPCG solver with the Jacobi and GLSPOLY preconditioners are significantly better.
-For problems where the MILU0 preconditioner is needed there is no real benefit to using the GPGPU because of the required additional CPU-GPGPU memory transfers -- for these cases parallel CPU solution would be adequate.
+\color{blue}The primary reason for the effectiveness of parallel GPGPU solutions is the number of cores available on GPGPUs relative to the number available on CPUs.
+\color{black}For problems where the MILU0 preconditioner is needed there is no real benefit to using the GPGPU because of the required additional CPU-GPGPU memory transfers -- for these cases parallel CPU solution would be adequate.
+\color{blue}Improved approaches for parallelization of incomplete factorization preconditioners is an active area of research (for example see, Li and Saad 2010; Naumov 2011; Hwang 2012).
+For example, Naumov (2011) was able to achieve an average speedup of 2 using a data-dependency graphing approach.
+Therefore, it is expected that additional research will improve the speedup possible with incomplete factorization preconditioners and ultimately the effectiveness of parallel solvers for MODFLOW and other numerical simulators.\color{black}
 
@@
 The benefits of using the GPGPU increases as the problem size increases.
+\color{blue}We suspect there are some problems that will require stronger preconditioners than the Jacobi and GLSPOLY preconditioners.
+For these problems we would suggest the parallel CPU solution with the MILU0 preconditioner.
+As a result, it seems use of hardware having multi-core CPUs and one or more GPGPUs is the most effective strategy for running the range of possible MODFLOW \color{cyan}[including MODFLOW variants such as MODFLOW-FMP2 (Schmid and Hanson 2009)] \color{blue} simulations in parallel.\color{black}
+
 In this study, problem size was limited by the RAM available on the GPGPU.
@@
 The UPCG solver is designed to solve  square, symmetric, positive-definite coefficient matrices.
-Modification of the UPCG solver to allow solution of nonsymmetric matrices, using the BiConjugate Gradient Stabilized method or other similar method (Barrett \textit{et al.} 1994), would permit use of the UPCG solver with MODFLOW-NWT (Niswonger \textit{et al.} 2011), MT3DMS (Zheng and Wang 1999), or SEAWAT (Langevin \textit{et al.} 2007).
+Modification of the UPCG solver to allow solution of nonsymmetric matrices, using the BiConjugate Gradient Stabilized method or other similar method (Barrett \textit{et al.} 1994), would permit use of the UPCG solver with MODFLOW-NWT (Niswonger \textit{et al.} 2011), MT3DMS (Zheng and Wang 1999), SEAWAT (Langevin \textit{et al.} 2007), \color{cyan} or the SWR1 Process for MODFLOW-2005 and MODFLOW-NWT (Hughes \textit{et al.} 2012)\color{black}.
 
@@
 It is possible that the software may need to be modified for application of the code on GPGPU models other than the Tesla C2050.
-And finally, python scripts used to create the MODFLOW datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
+And finally, UPCG solver input instructions and python scripts used to create the MODFLOW datasets evaluated using the standard PCG solver with the MIC preconditioner and the UPCG solver are also available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
 
@@
 
+\item \color{cyan} Harbaugh, A.W.
+1995.
+Direct solution package based on alternating diagonal ordering for the U.S.
+Geological Survey modular finite-difference ground-water flow model: U.S.
+Geological Survey Open-File Report 95-288, 46 p.\color{black}
+
 \item Harbaugh, A.W. 2005. MODFLOW-2005, the U.S. Geological Survey modular ground-water model -- the Ground-Water Flow Process. U.S. Geological Survey Techniques and Methods 6-A16, variously p.
@@
 
+\item \color{cyan} Hughes, J.D., Langevin, C.D., Chartier, K.L., White, J.T.
+2012.
+Documentation of the Surface-Water Routing (SWR1) Process for modeling surface-water flow with the U.S.
+Geological Survey Modular Ground-Water Model (MODFLOW–2005): U.S.
+Geological Survey Techniques and Methods, book 6, chap.
+A40 (Version 1.0), 113 p.\color{black}
+
+\item \color{blue} Hwang, H.T.
+2012.
+Development of a parallel computational framework to solve flow and transport in integrated surface-subsurface hydrologic systems.
+PhD dissertation, University of Waterloo, 133 p.\color{black}
+
+\item \color{cyan} Ji, Xiaohui, Cheng, Tangpei, Wang, Qun 2010.
+A simulation of large-scale groundwater flow on CUDA-enabled GPUs.
+In Proceedings of the 2010 ACM Symposium on Applied Computing (SAC '10).
+ACM, New York, NY, USA, 2402-2403.
+doi: 10.1145/1774088.1774588\color{black}
+
 \item Ji, Xiaohui, Cheng, Tangpei, Wang, Qun 2012. CUDA-based solver for large-scale groundwater flow simulation. Engineering with Computers 28, no. 1: 13-19. doi: 10.1007/s00366-011-0213-2
 
-\item Langevin, C.D., Thorne, D.T., Jr., Dausman, A.M., Sukop, M.C., and Guo, Weixing 2007.
+\item Langevin, C.D., Thorne, D.T., Jr., Dausman, A.M., Sukop, M.C., Guo, Weixing 2007.
 SEAWAT Version 4: A Computer Program for Simulation of Multi-Species Solute and Heat Transport.
@@
 Geological Survey Techniques and Methods Book 6, Chapter A22, 39 p.
+
+\item \color{cyan}Lawson, C.
+L., Hanson, R.
+J., Kincaid, D.
+R., Krogh, F.
+T.
+1979.
+Basic Linear Algebra Subprograms for Fortran Usage.
+ACM Transactions on Mathematical Software 5: no.
+3, 308-323.
+doi: 10.1145/355841.355847\color{black}
 
@@
 
+\item \color{blue} Naumov, Maxim 2011.
+Incomplete-LU and Cholesky preconditioned iterative methods using CUSPARSE and CUBLAS.
+NVIDIA technical report, 16 p.
+Available from \url{http://developer.download.nvidia.com/devzone//devcenter/cuda/files/psts_white_paper_final.pdf}\color{black}
+
 \item Niswonger, R.G., Panday, Sorab, and Ibaraki, Motomu 2011. MODFLOW-NWT, A Newton formulation for MODFLOW-2005. U.S. Geological Survey Techniques and Methods 6-A37, 44 p.
@@
 \item NVIDIA 2011. NVIDIA CUDA™: NVIDIA CUDA C Programming Guide Version 4.1.
+
+\item \color{cyan} NVIDIA 2012.
+CUDA Toolkit 4.2 -- CUBLAS Library: PG-05326-041\_v01, 99 p.
+Available from \url{http://developer.download.nvidia.com/compute/DevZone/docs/html/CUDALibraries/doc/CUBLAS_Library.pdf}  \color{black}
 
@@
 \item Scandrett, C. 1989. Comparison of several iterative techniques in the solution of symmetric banded equations on a two-pipe Cyber 250, Applied Mathematics and Computation 34: 95-112.
+
+\item \color{cyan}Schmid, Wolfgang and Hanson R.T.
+2009.
+The Farm Process Version 2 (FMP2) for MODFLOW-2005 - Modifications and Upgrades to FMP1: U.S.
+Geological Survey Techniques in Water Resources Investigations, Book 6, Chapter A32, 102p.\color{black}
 
@@
 \begin{table}[hp]
+\caption{Convergence summary for simulations using the UPCG solver with the Jacobi, MILU0, and GLSPOLY preconditioners, the PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner.
+The total number of outer and inner iterations are shown (outer/inner).
+Shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
+   \label{TableCPUITSummary}
+   \begin{tabular}{ l | c | c | c | c | c |}
+ 		\cline{2-6}
+		& \multicolumn{5}{c|}{Convergence History} \\ \cline{1-6}
+		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & Jacobi & MILU0 & GLSPOLY & MIC & GMG \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } & 5/955 & 6/136 & 6/130 & 6/140 & 5/7 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } & 6/2629 & 7/298 & 7/352 & 7/312 & 5/9 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 8/4624 & 8/547 & 7/765 & 9/574 & 6/10 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 21/17862 & 8/980 & 8/1752 & 8/1022 & 7/14 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & \cellcolor{gray}50/50000 & 9/1758 & 9/3257 & 9/1859 & 8/16 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 5/811 & 6/145 & 5/123 & 6/151 & 5/7 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } & 5/2073 & 6/315 & 5/276 & 6/334 & 7/12 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 6/4102 & 7/562 & 6/633 & 7/597 & 8/16 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 17/15342 & 7/864 & 5/1363 & 7/902 & 20/33 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } & 5/788 & 5/154 & 5/124 & 6/167 & 5/6 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } & 5/1967 & 5/319 & 5/261 & 5/335 & 6/11 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } & 6/4059 & 6/633 & 5/637 & 7/685 & 7/14 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } & 16/14848 & 6/1003 & 5/1359 & 6/1069 & 23/33 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 5/955 & 6/273 & 5/164 & 6/291 & 5/7 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 5/1908 & 5/512 & 5/275 & 5/546 & 6/9 \\ \cline{1-6}
+        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 6/3913 & 6/955 & 5/521 & 7/1014 & 7/13 \\ \cline{1-6}
+       \end{tabular}
+   \end{table}
+
+%Table2
+\newpage
+\begin{table}[hp]
    \caption{Speedup of UPCG solver GPGPU simulations relative to CPU simulations for the same UPCG preconditioner.
@@
 
-%Table2
+%Table3
 \newpage
@@
 
-%Table3
+%Table4
 \newpage
@@
         \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 0.835/0.968/0.928 & 1.720/1.734/1.651 & 0.815/0.798/0.914 \\ \cline{1-4}
-       \end{tabular}
-   \end{table}
-
-%Table4
-
-\newpage
-\begin{table}[hp]
-\caption{Convergence summary for simulations using the PCG solver with the MIC preconditioner, the GMG solver with the ILU0 preconditioner, and the UPCG solver with the Jacobi, MILU0, and GLSPOLY preconditioners.
-The total number of outer and inner iterations are shown (outer/inner).
-Shaded cells indicate simulations that did not converge within 50 outer iterations requiring as much as 1,000 inner iterations.}
-   \label{tab:UPCGITSummary}
-   \begin{tabular}{ l | c | c | c | c | c |}
- 		\cline{2-6}
-		& \multicolumn{5}{c|}{Convergence History} \\ \cline{1-6}
-		\multicolumn{1}{|c|}{Columns $\times$ Rows $\times$ Layers} & MIC & GMG & Jacobi & MILU0 & GLSPOLY \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 1 } & 6/140 & 5/7 & 5/955 & 6/136 & 6/130 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 1 } & 7/312 & 5/9 & 6/2629 & 7/298 & 7/352 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 1 } & 9/574 & 6/10 & 8/4624 & 8/547 & 7/765 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 1 } & 8/1022 & 7/14 & 21/17862 & 8/980 & 8/1752 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 4000 $\times$ 4000 $\times$ 1 } & 9/1859 & 8/16 & \cellcolor{gray}50/50000 & 9/1758 & 9/3257 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 2 } & 6/151 & 5/7 & 5/811 & 6/145 & 5/123 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 2 } & 6/334 & 7/12 & 5/2073 & 6/315 & 5/276 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 2 } & 7/597 & 8/16 & 6/4102 & 7/562 & 6/633 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 2 } & 7/902 & 20/33 & 17/15342 & 7/864 & 5/1363 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 3 } & 6/167 & 5/6 & 5/788 & 5/154 & 5/124 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 3 } & 5/335 & 6/11 & 5/1967 & 5/319 & 5/261 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 3 } & 7/685 & 7/14 & 6/4059 & 6/633 & 5/637 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 2000 $\times$ 2000 $\times$ 3 } & 6/1069 & 23/33 & 16/14848 & 6/1003 & 5/1359 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 200 $\times$ 200 $\times$ 10 } & 6/291 & 5/7 & 5/955 & 6/273 & 5/164 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 500 $\times$ 500 $\times$ 10 } & 5/546 & 6/9 & 5/1908 & 5/512 & 5/275 \\ \cline{1-6}
-        \multicolumn{1}{| l |}{ 1000 $\times$ 1000 $\times$ 10 } & 7/1014 & 7/13 & 6/3913 & 6/955 & 5/521 \\ \cline{1-6}
        \end{tabular}

### 2012-09-04 ResponseTo_GW20120518-0099_InternalReview.tex
Modifications to review response documents

@@
 Review comments are included (\textit{in italics}) prior to our responses.
-Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.
+Revised material in the manuscript are indicated with \color{cyan} cyan \color{black} text if it has not been addressed previously in the responses to journal reviewer comments (which are indicated with \color{blue} blue \color{black} text).
 
@@
 
-Response
+We have added a brief discussion of current research being conducted by others of improved parallelization of incomplete factorization preconditioners and there is a discussion of possible future enhancements of the UPCG solver in the last pragraph of the \textbf{Conclusions} section.
 
-\item \textit{It would be also nice if the authors can indicate where and how to obtain the source code (will it be in public domain just like the MODFLOW source code?.}
+We have not added a practical example because we believe the test problems presented represent a synthetic but realistic system and is a good control problem for evaluating the effectiveness of the UPCG solver -- and this is the ultimate goal of this paper.
 
-Response
+\item \textit{It would be also nice if the authors can indicate where and how to obtain the source code (will it be in public domain just like the MODFLOW source code?).}
+
+We have included a new section titled \textbf{How to Obtain the Software} that includes this information.
 
@@
 
-Response
+This will be included in the github repository that is described in the \textbf{How to Obtain the Software} section.
 
@@
 
-Response.
+Undoubtedly, there is room for improving the efficiency of the UPCG solver code (and probably any experimental code) but we believe the issues related to parallelization of the Jacobi, GLSPOLY, and MILU0 preconditioners would still apply even if the code were more efficient.
+However having said that, we believe there is some problem dependence to the optimal number of Picard and inner iterations for a given problem.
+We set up the test problem so that we maximized the amount of time spent in the linear solver to maximize the conjugates evaluated and allowed a large number of Picard iterations to ensure convergence was achieved for most simulations.
+
+See responses to inline comments in MODFLOW-GPU\_03.jdhResponse.pdf.
 
@@
 
-Response.
+We have included our best attempt at guidance based on our simulations for effectively using the various preconditioners and/or hardware options in the \textbf{Discussion} section.
 
@@
 
-Response.
+We haven't seen significant improvements with chip specific compiler switches, MKL libraries (beyond improvements related to OpenMP), data alignment, or vectorization (the UPCG solve is a vector solver that uses CRS indirect indexing).
+Co-arrays may be beneficial for cluster applications but probably not as beneficial for multi--core CPUs because of the limited number of available threads.
+A cluster solution is something that we are considering but would be less accessible to many readers of Ground Water.
+We suspect that parallel CPU and GPGPU solutions would be more accessible to readers.
 
@@
 The second reference (Ji \textit{et al.}, 2012) is a reference in the manuscript but this is difficult to determine from the link provided.
-The first reference (Ji \textit{et al.}, 2012) was not included in the original manuscript since it is a conference proceeding and may be difficult for some readers to locate.
+The first reference (Ji \textit{et al.}, 2010) was not included in the original manuscript since it is a conference proceeding and may be difficult for some readers to locate.
 It has been added to ensure Ji \textit{et al.}, 2010 is credited as being the first application of solving MODFLOW using GPGPUs.
+
+\item \textit{Define the term speedup in the paper.
+Speedup is used in many different ways: solver speedup, speedup of parallelized loops, or speedup of total simulation time.
+Since simulation efficiency is at the core of this paper, it will be useful to mention what the calculated speedups are based on.}
+
+The definition of speedup has been clarified to indicate that it is based solely on the time spent within solver routines.
+
+\item \textit{Are there any drawbacks to using GPUs and how these compare to using CPUs: limitations on the size of the problem; a general cost comparison (twice/thrice) etc.}
+
+Information on the RAM limitations of the specific GPGPUs is given in the \textbf{Test Cases} and \textbf{Conclusions} sections.
+A summary under which use of parallel GPGPU and CPU approaches should be beneficial has been added to the \textbf{Discussion} section.
+
+\item \textit{Page 9 mentions that two different machines were used (2.4 GHz and 3 GHz) for comparing the CPU and GPU simulations.
+Were comparisons made between the two machines by running exactly same simulation runs on the two CPUs (without multithreading and by using 2 cores) to benchmark whether there are inherent differences in the performance of the two machines?
+Besides the processor speed other factors like bus speed, motherboard, hard drives, etc.
+could be impacting the results.}
+
+Serial and parallel runs were made on both machines so that the correct speedup could be calculated on each machine.
+Additional text has been added to clarify the approach used for model simulations on both machines.
+
+\item \textit{Did the authors study if a relationship existed between the total number of inner iterations and the speedups in the test simulations?
+If such a comparison was made, it may be interesting to share that with the readers.
+That would be an indication to the users as to which problems may reap the benefits of GPU more than others.}
+
+Yes there is a relationship between inner iterations and speedup.
+The runtime information shown in \textbf{figure 4} tells part of the story but marginal to poor speedups for the Jacobi preconditioner is related to the number of iterations required for convergence.
+
+A new table has been added (\textbf{table 1} in the revised manuscript) that summarizes the number of Picard and inner iterations used by each solver and preconditioner evaluated.
+Additional text that discusses the relationship between the strength of the preconditioner and the observed speedup has been added to the \textbf{Discussion} section.
+We believe this will provide readers with some guidance on when to try parallel GPGPU or CPU approaches to a given problem.
+
+\item \textit{Was the total number of inner and outer iterations compared between the CPU and GPU simulations?
+The number should come out to be very similar in all simulations since the only thing GPU is doing is to parallelize the same computations that CPU is doing.
+This comparison would provide more confidence in the results that no other processes/algorithms besides GPU and its parallelization are responsible for the speedups.}
+
+A new table has been added (\textbf{table 1} in the revised manuscript) that summarizes the number of Picard and inner iterations used by each solver and preconditioner evaluated.
+Additional text has been added to the \textbf{Results} section to summarize the data included on the new table.
+
+\item \textit{There are more robust and faster solvers available besides PCG2 (XMD of MODFLOW‐NWT; PCG5 of MODFLOW‐SURFACT).
+It would be interesting to see how the GPU speedup (of 8 when compared to PCG2), compares with faster algorithms used in other solvers like XMD and PCG5.}
+
+There are certainly other solvers available that have been incorporated into MODFLOW or MODFLOW-variants by others but we have intentionally limited our discussion to standard solvers available for MODFLOW-2005.
+This study was meant to be an evaluation of the applicability of GPGPU parallelization to a PCG solver not all solvers.
+The findings of this study should be applicable to other solvers and could be implemented in a similar fashion.
+It is our expectation that speedups for the solvers mentioned would be similar to the UPCG solver with the MILU0 preconditioner because of the difficulty of parallelizing an incomplete factorization.
+
+It is our assessment that it is the parallel GLSPOLY preconditioner that makes use of the GPGPU useful for MODFLOW at this point.
+Furthermore, we expect that future research on ways to parallelize incomplete factorization preconditioners will greatly improve the speedup for parallel solvers on CPUs and GPGPUs.
+
+\item \textit{The LMG2 (SAMG) solver is a parallel solver that claims speedups of up to 100 times when compared to the PCG solver  (\url{http://www.scai.fraunhofer.de/fileadmin/download/samg/paper/modflow11.pdf}; \url{http://www.scai.fraunhofer.de/fileadmin/download/samg/paper/Modflow_Paper.pdf}).
+The authors may wish to comment about the combination of the GPU technology and the SAMG
+solver.}
+
+The authors have considered parallelization of a multi-grid solver but have not performed any evaluation of the effectiveness of parallelization of a multi-grid solver relative to a traditional Krylov solver.
+Parallel multi-grid solvers are reported to be more scalable than traditional Krylov solvers.
+To keep the paper on topic - comparison of parallel strategies on the UPCG solver with speedups based on the PCG solver - we have elected not to add a discussion of parallelization of the LMG2 solver on GPGPUs.
+
+\item \textit{How does PCG and UPCG sequential simulation compare?}
+
+PCG and UPCG are competitive.
+For some problems PCG is slightly faster for other problem UPCG is slightly faster.
+Iteration information is summarized on \textbf{table 1} in the revised manuscript.
+In general, the performance of PCG suffers with larger problems and is summarized in the \textbf{Results} section.
+
+\item \textit{Mention that the groundwater simulation results are the same irrespective of the solver/processor used.}
+
+This has been added to the end of the \textbf{Problem Description} subsection.
+
+\item \textit{Line 35: replace SOR with SSOR.}
+
+Corrected
+
+\item \textit{Line 40: also refer to other more efficient solvers, including the XMD, LMG2, and PCG5 solvers.}
+
+There are certainly other solvers available that have been incorporated into MODFLOW or MODFLOW-variants by others but we have intentionally limited our discussion to standard solvers available for MODFLOW-2005.
+XMD has been incorporated into MODFLOW-NWT but it's real strength is its ability to solve non-symmetric matrices using more sophisticated level-fill and drop tolerance approaches.
+Since the test problems we are evaluating are symmetric it is questionable whether these approaches would add convergence improvements that outweigh the additional numerical overhead of these approaches.
+In my experience XMD is less efficient in simulating standard MODFLOW problems than the PCG and UPCG solvers.
+
+\item \textit{Line 44: it is important to note that speedup in this paper is based on ``solver computing time'' as defined in the paper.
+Upon testing, I had found that the loops parallelized took 25--30\% of the total simulation time for the problem I was testing.
+Effectively, 5 times speedup (20\%) of 25--30\% comes out to 5--6\% saving in the total simulation time.}
+
+This clarification has been added.
+
+\item \textit{Line 92: “a” appears twice.}
+
+Corrected -- moved to \textbf{box 2}.
+
+\item \textit{Lines 161 and 163: is there a citation for CUBLAS and BLAS? If these are acronyms, what do they stand for?}
+
+The acronyms has been defined in the revised text. References for CUBLAS and BLAS have been added.
+
+\item \textit{Line 162: Saad, 2003 is misspelled.}
+
+Corrected
+
+\item \textit{Line 173: what is GEMV? Is this an acronym?}
+
+It is an abbreviation for general matrix-vector (GEMV) product. This has been added to the manuscript.
+
+\item \textit{Lines 193 and 196: Xeon is misspelled.}
+
+Corrected
+
+\item \textit{Lines 251--252: is this treatment of RCLOSE applied to PCG solver as well?
+If not, then the comparison between PCG and UPCG is unfair since the two solvers would converge to different RCLOSE values.}
+
+Yes. This has been added.
+
+\item \textit{Line 317: if MIC is more efficient, was parallelization of the MIC preconditioner attempted? Please make a statement as to why not.}
+
+We did not intend to say anything more than the standard PCG solver is generally fairly efficient rather than something specific about the efficiency of the MIC preconditioner relative to the Jacobi, GLSPOLY, or MILU0 preconditioners available with the UPCG solver.
+This sentence has been revised to clarify this point.
+
+\item \textit{Table 3: several values in this table are less than 1, indicating that UPCG is slower than PCG. What is the possible reason of this slowdown?}
+
+Additional clarification has been provided for PCG(MIC)/UPCG-OpenMP speedups less than 1 in the \textbf{Discussion} section.
 

### 2012-09-04 ResponseTo_GW20120518-0099_Review.tex
Modifications to review response documents

@@
 \noindent
-\textbf{Add some introductory text about how reviewer comments were helpful and have been addressed in the revised manuscript}
+We greatly appreciate the comments provided by the three reviewers and the editor.
+We have addressed all of the comments in this document.
+In general, we have made all of the suggested modifications.
+In a few cases we disagreed with reviewer comments but have provided a justification for not accepting the comment.
 
-Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.
+Reviewer comments are included (\textit{in italics}) prior to our responses.
+Revised material in the manuscript is indicated using \color{blue} blue \color{black}text.
+Text indicated in \color{cyan} cyan \color{black} in the revised manuscript represent changes in response to internal USGS reviewers.
 
@@
 
-\item \textit{Thorough comparisons were made for different preconditioners and different parallelization strategies and the authors did a great job of making the comparisons fair.
+\item \label{kpropertydiscussion} \textit{Thorough comparisons were made for different preconditioners and different parallelization strategies and the authors did a great job of making the comparisons fair.
 Considering the results of the various preconditioners, it appears that Jacobi preconditioning is good enough for many of the problems, which leads me to believe that the K fields are fairly smooth (and the spread of the eigenvalues are nicely clustered).
@@
 
-Response.
+Evaluation of the maximum and minimum eigenvalues for the original 1000 $\times$ 1000 problem suggests that there is clustering of eigenvalues in the original $\mathbf{A}$ matrix.
+It was not possible to evaluate the full spread of the eigenvalues because of the size of the $\mathbf{A}$ matrix.
+Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner, however, indicates there are two clusters with the first ranging from approximately 2.5$\times 10^{-4}$ to 1$\times 10^{-2}$ and approximately 1.5$\times 10^{-2}$.
+
+To look at the effect that eigenvalue clustering has on the effectiveness of the Jacobi preconditioner we have evaluated a number of alternative problems in an effort to reduce the smoothness of the fields.
+The additional fields that were evaluated are based on the fields used to evaluate the performance of the UPCG solver presented in the manuscript and included: (1) random permutations values in the original fields; (2) fields that used the average hydraulic conductivity in select quantiles of the original fields (to develop a zoned field); (3) a constant field based on the mean hydraulic conductivity of the original fields; and (4) a stochastic field that used the same properties of the original fields except with a variance that varied over 5 orders of magnitude (the original varied over 3 orders of magnitude).
+Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner indicates that the eigenvalues for (1) are clustered in three groups with a cluster covering a large range from approximately 5$\times 10^{-4}$ to 7.5$\times 10^{-3}$, (2) show no visible signs of clustering, (3) show no visible signs of clustering, and (4) are clustered into two groups like the original field but span a larger range.
+
+For all of these alternative fields, the Jacobi preconditioner was able to sufficient to solve the problem and required a comparable number of iterations to solve the problem (approximately 4,500 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem) except for the fields with a larger variance (approximately 5,200 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem).
+The authors suspect the Jacobi preconditioner is sufficient in almost every case evaluated because of the number of iterations that are allowed (50,000).
+In fact, we were able to get the 4,000 $\times$ 4,000 $\times$ 1 problem to solve if the maximum number of iterations was increased (a solution was achieved in XX,XXX iterations).
+
+We have not included results for any of these additional simulations but have added additional discussion of how speedup may be affected by the hydraulic conductivity field has been added to the \textbf{Discussion} and \textbf{Conclusions} sections.
 
@@
 
-Response.
+OpenMP was not evaluated on PCG2 but was considered using the UPCG solver.
+We believe we can accurately report on conditions under which use of a GPGPU solution would be preferred to an OpenMP solution and vice-versa based on (1) the similarity of the effectiveness of the UPCG solver with the MILU0 preconditioner and the PCG solver with the MIC preconditioner and (2) the fact that UPCG solver speedups with OpenMP are comparable to the Intel compiler results reported in Dong and Li (2009) .
+We have revised the \textbf{Conclusions} section (see response to \textbf{Reviewer 2 Comment~\ref{kpropertydiscussion}}).
+We have also evaluated the effect that Intel compiler optimization has on OpenMP speedups and have revised the \textbf{Discussion} section to reflect the observation that speedups were comparable with and without use of the -O3 compiler switch.
+
+We have also included our assessment of the best approach for being able to run a range of MODFLOW problems in parallel has been added to the \textbf{Conclusions} section.
 
@@
 
-Response.
+See response to \textbf{Reviewer 2 Comment~\ref{kpropertydiscussion}}.
 
@@
 
-Response.
+We are aware of this work but have not attempted to implement it.
+One of the reasons we did not devote any time to its implementation is Naumov (2011) only reported an average speedup of 2.0 which is less than was observed with the GLSPOLY preconditioner.
+The work of Li and Saad (2010) pointed us in the direction of the GLSPOLY preconditioner.
+However, we have included a few sentences discussing recent work devoted to improving the effectiveness of incomplete factorization preconditioners in the \textbf{Conclusions} section.
 
@@
 
-\textbf{Table 4} has been added which includes information on the number of iterations required for each of the solvers and various preconditioners.
+A table has been added (revised \textbf{Table 1}) which includes information on the number of iterations required for each of the solvers and various preconditioners.
 Additional discussion has been added in the \textbf{Discussion} section to provide some guidance on the strength of the preconditioners for the problems evaluated.
@@
 
-Yes.
+Yes. See \textbf{equation 2-2} in Harbaugh (2005).
 
@@
 
-Yes.
+Yes. See \textbf{equation 2-9} in Harbaugh (2005).
 

### 2012-09-04 ResponseTo_GW20120518-0099_Review.tex
jwhite review

@@
 
-For all of these alternative fields, the Jacobi preconditioner was able to sufficient to solve the problem and required a comparable number of iterations to solve the problem (approximately 4,500 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem) except for the fields with a larger variance (approximately 5,200 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem).
+For all of these alternative fields, the Jacobi preconditioner was able to solve the problem and required a comparable number of iterations (approximately 4,500 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem) except for the fields with a larger variance (approximately 5,200 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem).
 The authors suspect the Jacobi preconditioner is sufficient in almost every case evaluated because of the number of iterations that are allowed (50,000).

### 2012-09-04 GW20120518-0099-r01.tex
jwhite review

@@
 For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem the speedup would be 2, 3, 4, and 5, respectively.
-However, ideal parallelization is rarely, if ever, achieved because of communication bottlenecks which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
+However, ideal parallelization is rarely, if ever, achieved because of communication overhead which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
 \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}
@@
 Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
-\color{cyan}Simulated water-levels for simulations that satisfy specified convergence criteria and use different (1) solvers, (2) preconditioners, and/or (3) hardware are within $\pm$RCLOSE m.\color{black}
+\color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$RCLOSE m.\color{black}
 

### 2012-09-04 ResponseTo_GW20120518-0099_Review.tex
Additional modifications

@@
 
-\item \textit{Issues with GPGPU processing for different schemes are scattered throughout the manuscript.
+\item \textit{Issues with GPGPU processing for different schemes are scattered throughout the man\-uscript.
 These issues should be collated into a separate section or a table.
@@
 The authors suspect the Jacobi preconditioner is sufficient in almost every case evaluated because of the number of iterations that are allowed (50,000).
-In fact, we were able to get the 4,000 $\times$ 4,000 $\times$ 1 problem to solve if the maximum number of iterations was increased (a solution was achieved in XX,XXX iterations).
+In fact, we were able to get the 4,000 $\times$ 4,000 $\times$ 1 problem to solve if the maximum number of iterations was increased (a solution was achieved in 66,861 iterations).
 

### 2012-09-05 ResponseTo_GW20120518-0099_Review.tex
Final modifictions prior to next review

@@
 
-The authors disagree with this comment.
-We believe the issues related to solution of the groundwater flow equation on a GPGPU are presented in the relevant locations and important issues are summarize again in the ``\textbf{Conclusions}'' section.
+We have reorganized some sections of the document.
+For other sections, however, we have retained the original discussion of ``issues'' because we believe the original location is relevant to the section and movement to a separate section would adversely affect the readability of the manuscript.
 

### 2012-09-05 GW20120518-0099-r01.tex
Final modifictions prior to next review

@@
 
-\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs.
-Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7.
+\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
+Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 \color{cyan}using 32 CPUs\color{black}.
 In addition to OpenMP and MPI, GPGPUs also \color{blue}present \color{black}an option for linear solver parallelization.
@@
 The main code is a sequential host code which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.
-\color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have to potential to spawn thousands to tens-of-thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
+\color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens-of-thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
 The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.
@@
 
-MODFLOW uses a cell-centered, finite-difference approximation of \textbf{equation~\ref{gwpde}} to solve for three-dimensional groundwater flow (Harbaugh 2005).
+MODFLOW uses a CCFD approximation of \textbf{equation~\ref{gwpde}} to solve for three-dimensional groundwater flow (Harbaugh 2005).
 Development of the finite-difference form of \textbf{equation~\ref{gwpde}} using the continuity equation for a cell results in
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number matrix-vector product and a constant times a vector plus a vector (axpy) BLAS operations \color{black} (Liang \textit{et al.} 2002; Saad 2003).
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of  BLAS operations (Liang \textit{et al.} 2002; Saad 2003); the BLAS operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
 
@@
 
-\color{blue}To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes if discretized using different horizontal and vertical cell sizes.
-The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear as a result of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$.
+\color{blue}To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes.
+The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear because of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$.
 Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.\color{black}
@@
 The head-dependent boundaries cause an ambient groundwater flow from left to right.
-Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d \color{cyan}and represents 46\% of the flow entering the model\color{black}.
+Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d.
 For multi-layer simulations, the pumping wells withdraw water from the lower model layer.
-There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries.
+There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries \color{cyan}and represents 46\% of the flow entering the model\color{black}.
 Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.
@@
 \subsection*{Results}
-\color{blue}The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and the combination of the GPGPU and CPU were identical to the number required on the CPU.
+\color{blue}The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and hybrid CPU/GPGPU approaches were identical to the number required on the CPU.
 The number of Picard and inner iterations required by the MILU0 preconditioner is comparable to the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
-The number of Picard iterations is also comparable between the MILU0, PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner; the GMG inner iterations are not comparable because of the differences in the inner iteration approaches of both the UPCG and PCG solvers and the GMG solver.
-The number of Picard and inner iterations required by the GLSPOLY preconditioner generally exceeds the MILU0 preconditioner and the PCG solver with the MIC preconditioner by a factor of but it is notably better the Jacobi preconditioner.
+The number of Picard iterations is also comparable between the MILU0 preconditioner, PCG solver with the MIC preconditioner, and the GMG solver with the ILU0 preconditioner; the GMG inner iterations are not comparable because of differences in the inner iteration approaches of both the UPCG and PCG solvers and the GMG solver.
+The number of Picard and inner iterations required by the GLSPOLY preconditioner is generally greater than the MILU0 preconditioner and the PCG solver with the MIC preconditioner but it is notably better the Jacobi preconditioner.
 In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.\color{black}
@@
 
-PCG(MIC)/GPGPU speedups for the MILU0 preconditioner are comparable to CPU/ GPGPU speedups; the similarity of PCG(MIC)/GPGPU and CPU/GPGPU speedups for the MILU0 preconditioner are a result of the similarity in time required to perform forward and backward substitutions during application of the preconditioner with the PCG and UPCG solvers, as the MIC is similar to MILU0 .
+PCG(MIC)/GPGPU speedups for the MILU0 preconditioner are comparable to CPU/ GPGPU speedups; the similarity of PCG(MIC)/GPGPU and CPU/GPGPU speedups for the MILU0 preconditioner are a result of the similarity in time required to perform forward and backward substitutions during application of the preconditioner with the PCG and UPCG solvers, as the MIC is similar to MILU0.
 PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners range from less than 1.0 to more than 8.0, which are notably less than CPU/GPGPU speedups.
@@
 The relatively small PCG(MIC)/UPCG-OpenMP speedups observed with the UPCG solver compiled using the Intel Fortran compiler is consistent with the observations of Dong and Li (2009).
-\color{blue} We observed similar PCG(MIC)/UPCG-OpenMP speedups with and without optimization of the executable compiled by the Intel Fortran compiler.\color{black}
+\color{blue} We observed similar PCG(MIC)/UPCG-OpenMP speedups with and without optimization of the executable compiled using the Intel Fortran compiler.\color{black}
 
@@
 
-The PCG algorithm was implemented for use on GPGPUs using the NVIDIA \color{blue}CUBLAS library\color{black}.
+The PCG algorithm was implemented for use on GPGPUs using the NVIDIA \color{blue}CUDA API\color{black}.
 The availability of this API makes it relatively simple to develop parallel GPGPU code.
@@
 The master thread executes sequential operations and initiates thread forks used in parallel regions.
+Adapted from Dong and Li (2009).
 (B) GPGPU architecture showing the relation of streaming multiprocessors and CUDA cores.

### 2012-09-16 GW20120518-0099-r01.tex
Modified document

@@
 Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
-\color{blue}Increase model runtimes are typically a direct result of increases in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
+\color{blue}Increase model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
 \color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
@@
 \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem.
-For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem the speedup would be 2, 3, 4, and 5, respectively.
+For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively.
 However, ideal parallelization is rarely, if ever, achieved because of communication overhead which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
@@
 
-\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
+\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
 Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 \color{cyan}using 32 CPUs\color{black}.
-In addition to OpenMP and MPI, GPGPUs also \color{blue}present \color{black}an option for linear solver parallelization.
-In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and as a result are capable of executing more floating point operations per second.
-\color{black}Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
+In addition to OpenMP and MPI, \color{cyan}general-purpose graphics processing units (GPGPUs) \color{black}also \color{blue}present \color{black}an option for linear solver parallelization.
+In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second.
+\color{black}Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the Compute Unified Device Architecture (CUDA) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
 For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems.
@@
 
-In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured UPCG linear solver for a unconfined groundwater flow problem.
+In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured  preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem.
 We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete LU (MILU0), and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU.
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of  BLAS operations (Liang \textit{et al.} 2002; Saad 2003); the BLAS operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of independent operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
 
@@
 
-Available CUBLAS matrix-vector products and level-1 BLAS operations were used to parallelize the GPGPU capabilities of the UPCG linear solver.
+Available \color{cyan}CUBLAS-enabled BLAS operations \color{black}were used to parallelize the GPGPU capabilities of the UPCG linear solver.
 Because of the poor performance of forward and back substitution operations used by the ILU0 and MILU0 preconditions observed during initial testing of the UPCG linear solver and observed by others (Li and Saad 2010), application of these preconditioners is performed on the CPU, resulting in a hybrid CPU/GPGPU solver.
@@
 
-For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and level-1 BLAS operations.
+For parallel CPU simulations, OpenMP directives were added to all matrix-vector and vector operations.
 The OpenMP matrix-vector product was constructed using a level-2 BLAS \color{cyan}general matrix-vector (GEMV) product \color{black}approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
 OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively.
-The number of OpenMP threads used for matrix-vector products and level-1 BLAS operations is specified separately in the UPCG input file.
-For sequential CPU simulations,  separate matrix-vector product and level-1 BLAS operation routines that exclude OpenMP directives are used to eliminate all OpenMP overhead.
+The number of OpenMP threads used for matrix-vector products and vector operations is specified separately in the UPCG input file.
+For sequential CPU simulations, separate matrix-vector product and vector operation routines that exclude OpenMP directives are used to eliminate all OpenMP overhead.
 
@@
 
-The amount of time applying the MILU0 and GLSPOLY preconditioners, performing matrix-vector operations, level-1 BLAS operations, and transferring data between the CPU and GPU for a select number of simulations are shown graphically in \textbf{figure~\ref{FigBLASOperationResults}}.
+The amount of time applying the MILU0 and GLSPOLY preconditioners, performing matrix-vector operations, \texttt{axpy}, all vector operations (\textit{i.e.}, dot-product), and transferring data between the CPU and GPU for a select number of simulations are shown graphically in \textbf{figure~\ref{FigBLASOperationResults}}.
 Matrix-vector operations account for approximately half of the time spent in the UPCG solver for sequential CPU simulations using either the MILU0 or GLSPOLY preconditioners.
@@
 This is because the application of the MILU0 preconditioner is always performed on the CPU.
-For the MILU0 preconditioner, all of the speedup occurs during the conjugate gradient routine in the level-1 BLAS operations [matrix-vector multiplication and vector operations (dot products)], which are carried out on the GPGPU after application of the preconditioner.
+For the MILU0 preconditioner, all of the speedup occurs during the conjugate gradient routine in the BLAS operations [matrix-vector multiplication and vector operations (dot-products)], which are carried out on the GPGPU after application of the preconditioner.
 Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases \color{cyan}(\textbf{fig.~\ref{FigBLASOperationResults}a} and \textbf{b})\color{black}.
@@
 When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions.
-The kernel executed by the UPCG solver on the GPGPU is the preconditioned conjugate gradient algorithm.\\ \color{black}}}
+The kernel functions executed by the UPCG solver on the GPGPU are the BLAS operations in the preconditioned conjugate gradient algorithm (\textbf{algorithm~\ref{pcgalg}}).\\ \color{black}}}
 \bigskip
@@
 \parbox{5.75in}{
-\color{blue}\textbf{\\Box 2\\NVIDIA GPGPU Architecture}
+\color{blue}\textbf{\\Box 2\\NVIDIA GPGPU Architecture and Kernel Execution}
 
@@
 
-\hspace{15pt}A CUDA kernel function is executed on a CUDA grid, which is a grouping of threads.
-Each CUDA grid is composed of one or more blocks that are executed on each CC.
+\hspace{15pt}The operations in a CUDA kernel function are distributed as evenly as possible across a CUDA grid, which is a grouping of threads.
+Each CUDA grid is composed of one or more blocks that execute the kernel function on each CC.
 For increased throughput, threads within a block are further split into warps, which is the maximum number of threads that are managed and processed simultaneously on each CC.
 Multiple warps execute one after another to complete block processing.
-All CCs within a SM share a single instruction, which results in massively parallel processing potential.
+All CCs within a SM execute a single kernel at at time, which results in massively parallel processing potential.
 GPGPU architecture is shown graphically in  \textbf{figure~\ref{FigParallel}}.
@@
 	\end{subfigure}
- 	\caption{Time, in seconds, spent applying the preconditioner, performing BLAS level 1 operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
+ 	\caption{Time, in seconds, spent applying the preconditioner, performing BLAS operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
 Note operation execution times have been normalized using the maximum preconditioner application time for each problem to facilitate comparison of different model runs.}

### 2012-09-16 GW20120518-0099-r01.tex
changes to doc

@@
 Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
-\color{blue}Increase model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
+\color{blue}Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
 \color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
@@
 
-\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
+\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
 Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 \color{cyan}using 32 CPUs\color{black}.
-In addition to OpenMP and MPI, \color{cyan}general-purpose graphics processing units (GPGPUs) \color{black}also \color{blue}present \color{black}an option for linear solver parallelization.
+In addition to OpenMP and MPI, general-purpose graphics processing units (GPGPUs) also \color{blue}present \color{black}an option for linear solver parallelization.
 In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second.
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of independent operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
 
@@
 
-Available \color{cyan}CUBLAS-enabled BLAS operations \color{black}were used to parallelize the GPGPU capabilities of the UPCG linear solver.
+Available CUBLAS matrix-vector products and vector operations were used to parallelize the GPGPU capabilities of the UPCG linear solver.
 Because of the poor performance of forward and back substitution operations used by the ILU0 and MILU0 preconditions observed during initial testing of the UPCG linear solver and observed by others (Li and Saad 2010), application of these preconditioners is performed on the CPU, resulting in a hybrid CPU/GPGPU solver.
@@
 
-For parallel CPU simulations, OpenMP directives were added to all matrix-vector and vector operations.
+For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and vector operations.
 The OpenMP matrix-vector product was constructed using a level-2 BLAS \color{cyan}general matrix-vector (GEMV) product \color{black}approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
@@
 
-The amount of time applying the MILU0 and GLSPOLY preconditioners, performing matrix-vector operations, \texttt{axpy}, all vector operations (\textit{i.e.}, dot-product), and transferring data between the CPU and GPU for a select number of simulations are shown graphically in \textbf{figure~\ref{FigBLASOperationResults}}.
+The amount of time applying the MILU0 and GLSPOLY preconditioners, calculating matrix-vector products, \texttt{axpy}, all other vector operations (\textit{e.g.}, dot-product), and transferring data between the CPU and GPU for a select number of simulations are shown graphically in \textbf{figure~\ref{FigBLASOperationResults}}.
 Matrix-vector operations account for approximately half of the time spent in the UPCG solver for sequential CPU simulations using either the MILU0 or GLSPOLY preconditioners.
@@
 When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions.
-The kernel functions executed by the UPCG solver on the GPGPU are the BLAS operations in the preconditioned conjugate gradient algorithm (\textbf{algorithm~\ref{pcgalg}}).\\ \color{black}}}
+The kernels executed by the UPCG solver on the GPGPU are the CUDA-enable BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
 \bigskip
@@
 
-\hspace{15pt}The operations in a CUDA kernel function are distributed as evenly as possible across a CUDA grid, which is a grouping of threads.
-Each CUDA grid is composed of one or more blocks that execute the kernel function on each CC.
+\hspace{15pt}The total number of threads required to is a CUDA grid.
+A CUDA grid is composed of one or more blocks that are executed on one or more SM.
+All threads in a block run on the same SM with a dimension equal to the total number of times the kernel is applied.
+Each CUDA grid is .
 For increased throughput, threads within a block are further split into warps, which is the maximum number of threads that are managed and processed simultaneously on each CC.
-Multiple warps execute one after another to complete block processing.
-All CCs within a SM execute a single kernel at at time, which results in massively parallel processing potential.
-GPGPU architecture is shown graphically in  \textbf{figure~\ref{FigParallel}}.
+Multiple warps execute one after another to complete block processing of the kernel.
+All CCs within a SM execute the kernel concurrently, which results in massively parallel processing potential.
+GPGPU architecture and kernel execution are shown graphically in \textbf{figure~\ref{FigParallel}}.
 
@@
 	\end{subfigure}
- 	\caption{Time, in seconds, spent applying the preconditioner, performing BLAS operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
+ 	\caption{Time, in seconds, spent applying the preconditioner, calculating matrix-vector products, calculating \texttt{axpy}, calculating all other vector operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
 Note operation execution times have been normalized using the maximum preconditioner application time for each problem to facilitate comparison of different model runs.}

### 2012-09-16 GW20120518-0099-r01.tex
Revisions to document based on cost center review

@@
 When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions.
-The kernels executed by the UPCG solver on the GPGPU are the CUDA-enable BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
+The kernels executed by the UPCG solver on the GPGPU are the CUDA-enabled BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
 \bigskip
@@
 
-A GPGPU is a set of streaming multiprocessors (SM), each having many CUDA cores (CC); each CC is capable of executing many threads simultaneously.
+A GPGPU is a set of streaming multiprocessors (SM), each having many CUDA cores (CC); each SM is capable of executing many threads simultaneously.
 This nested architecture gives rises to the massively parallel capability of GPGPUs.
 
-\hspace{15pt}The total number of threads required to is a CUDA grid.
-A CUDA grid is composed of one or more blocks that are executed on one or more SM.
-All threads in a block run on the same SM with a dimension equal to the total number of times the kernel is applied.
-Each CUDA grid is .
-For increased throughput, threads within a block are further split into warps, which is the maximum number of threads that are managed and processed simultaneously on each CC.
+\hspace{15pt}Each kernel launches a grid of threads composed of one or more blocks of grouped threads.
+The number of blocks in a grid is determined by the size of the kernel data being processed or the number of SM.
+The total number of threads in a grid is equal to the number of threads per block times the number of blocks.
+CCs on each SM can run one or more blocks but blocks are restricted to a single CC.
+Threads within a block are further grouped into warps, which is the maximum number of threads that are managed and processed simultaneously in each block.
 Multiple warps execute one after another to complete block processing of the kernel.
-All CCs within a SM execute the kernel concurrently, which results in massively parallel processing potential.
 GPGPU architecture and kernel execution are shown graphically in \textbf{figure~\ref{FigParallel}}.

### 2012-09-17 GW20120518-0099-r01.tex
More box massaging

@@
 
-\hspace{15pt}Each kernel launches a grid of threads composed of one or more blocks of grouped threads.
+\hspace{15pt}Each kernel launches a CUDA grid, which is composed of one or more blocks of grouped threads.
+Threads within the same block can share certain types of physical memory on the GPGPU.
 The number of blocks in a grid is determined by the size of the kernel data being processed or the number of SM.
 The total number of threads in a grid is equal to the number of threads per block times the number of blocks.
-CCs on each SM can run one or more blocks but blocks are restricted to a single CC.
-Threads within a block are further grouped into warps, which is the maximum number of threads that are managed and processed simultaneously in each block.
+CCs on each SM can run one or more blocks but blocks are allocated to a single CC.
+Threads within a block are further subdivided into warps for processing; a warp is the maximum number of threads that are managed and processed simultaneously within each block.
 Multiple warps execute one after another to complete block processing of the kernel.
@@
 
-\hspace{15pt}The Tesla C2050 used in this study has the following specifications: SMs = 14, CC per SM = 32, threads per block = 1,024, and warp size = 32.
+\hspace{15pt}The Tesla C2050 used in this study has the following specifications: SMs = 14, CC per SM = 32, threads per block = 1,024, and threads per warp = 32.
 The multi-threaded instruction unit on each of the 14 Tesla C2050 SMs can run 48 warps concurrently, yielding a total of 21,504 concurrent threads (14 SMs $\times$ 48 warps per SM $\times$ 32 threads per warp).\\ \color{black}

### 2012-09-22 ResponseTo_GW20120518-0099_Review.tex
Final revisions prior for resubmittal to the journal

@@
 \noindent
-\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\
+\textbf{Use of general purpose graphics processing units with MODFLOW \\ (Ground Water Manuscript GW20120518-0099) }\\
 
@@
 \noindent
-We greatly appreciate the comments provided by the three reviewers and the editor.
+We greatly appreciate the comments provided by the three journal reviewers and the editor.
 We have addressed all of the comments in this document.
 In general, we have made all of the suggested modifications.
-In a few cases we disagreed with reviewer comments but have provided a justification for not accepting the comment.
+In a few cases, we disagreed with reviewer comments but have provided justification for all comments that were not accepted.
+
+We have moved a significant amount of the GPGPU jargon to two sidebar boxes based on the editor's suggestion.
+We believe moving the jargon to sidebar boxes makes the manuscript more readable.
+The peer reviewer comment's were outstanding and if possible we would like to acknowledge them if they are willing to be identified.
 
 Reviewer comments are included (\textit{in italics}) prior to our responses.
-Revised material in the manuscript is indicated using \color{blue} blue \color{black}text.
-Text indicated in \color{cyan} cyan \color{black} in the revised manuscript represent changes in response to internal USGS reviewers.
+Material revised based on journal reviewer comments are indicated using \color{blue} blue \color{black}text in the revised manuscript.
+\color{cyan} cyan \color{black}text in the revised manuscript represent changes made in response to internal USGS reviewers.
 
@@
 
-\item \textit{Line 26 - "can be *easily* parallelize"}
+\item \textit{Line 26 - ``can be *easily* parallelize"}
 
@@
 \item \textit{Lines 28 - 34 - there is some time discrepancies here from 2005 to 1984, to 1988.
-Perhaps just say "MODFLOW" on line 28 and remove the "since 1984" and insert "numerous" before "two-and" }
+Perhaps just say "MODFLOW" on line 28 and remove the ``since 1984" and insert ``numerous" before ``two-and" }
 
@@
 
-\item \textit{Line 47 "present" (plural)}
+\item \textit{Line 47 ``present" (plural)}
 
@@
 
-\item \textit{Line 351 - "developed to *use* the PCG" (one does not "solve" the algorithm) }
+\item \textit{Line 351 - ``developed to *use* the PCG" (one does not "solve" the algorithm) }
 
@@
 
-\item \textit{Line 236: "(10-0)/1000"is not necessary.}
+\item \textit{Line 236: ``(10-0)/1000"is not necessary.}
 

### 2012-09-22 GW20120518-0099-r01.tex
Last revisions prior to submittal for BAO approval
Final revisions prior for resubmittal to the journal

@@
 
-\title{Use of general purpose graphical processing units with MODFLOW}
-\author{Hughes, J.D.\\U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jdhughes@usgs.gov} \and White, J.T. \\ U.S. Geological Survey\\ Tampa, Florida\\ \texttt{jwhite@usgs.gov}}
-%\date{} % Activate to display a given date or no date (if empty),
-         % otherwise the current date is printed
+\title{Use of general purpose graphics processing units with MODFLOW}
+\author{Hughes, J.D. \and White, J.T.}
+\date{}
 
@@
 To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed.
-The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete LU factorization, and generalized least-squares polynomial preconditioners.
+The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete \color{cyan}lower-upper (LU) \color{black}factorization, and generalized least-squares polynomial preconditioners.
 The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP.
@@
 The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells.
-Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW PCG solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
+Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW \color{cyan}preconditioned conjugate gradient (PCG) \color{black}solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
 Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
@@
 The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver.
-These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG2) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
+These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
 
@@
 For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively.
-However, ideal parallelization is rarely, if ever, achieved because of communication overhead which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
+However, ideal parallelization is rarely, if ever, achieved because of communication overhead, which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
 \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}
@@
 In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem.
-We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete LU (MILU0), and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU.
+We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  \color{cyan}lower-upper (MILU0) factorization \color{black}, and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU.
 And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.
@@
 For GPGPU code implementations, the concept is similar to OpenMP.
-The main code is a sequential host code which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.
-\color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens-of-thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
+The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.
+\color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
 The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.
@@
 \noindent
-where  $K_{xx}$, $K_{yy}$, $K_{zz}$ are the hydraulic conductivity [L/T] along the x, y, and z-coordinate axes which are assumed to be parallel to the major axes of hydraulic conductivity; $h$ is the groundwater head [L]; $W$ is a volumetric flux per unit volume [L$^{3}$/L$^{3}$T] representing sources/sinks of water; S$_{S}$ is the specific storage [L$^{-1}$]; and t is time [T].
+where  $K_{xx}$, $K_{yy}$, $K_{zz}$ are the hydraulic conductivity [L/T] along the $x$, $y$, and $z$-coordinate axes, which are assumed to be parallel to the major axes of hydraulic conductivity; $h$ is the groundwater head [L]; $W$ is a volumetric flux per unit volume [L$^{3}$/L$^{3}$T] representing sources/sinks of water; S$_{S}$ is the specific storage [L$^{-1}$]; and $t$ is time [T].
 
@@
 The conjugate gradient (CG) iterative method is an efficient method for solving large systems of linear equations having a square, symmetric, positive-definite coefficient matrix.
-Pseudocode for the preconditioned conjugate gradient method is given in \textbf{algorithm~\ref{pcgalg}}.
+Pseudocode for the PCG method is given in \textbf{algorithm~\ref{pcgalg}}.
 
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and very easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
 
 \subsection*{Numerical Implementation}
-The UPCG linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations.
+The \color{cyan}unstructured preconditioned conjugate gradient (UPCG) \color{black}linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations.
 Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.
 
-Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
+Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row (CSR) storage format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
 This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012).
@@
 \subsection*{Problem Description}
-GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a 1,000 m square heterogeneous unconfined aquifer.
+GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a \color{cyan}1,000 m $\times$ 1,000 m \color{black} heterogeneous unconfined aquifer.
 The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively.
@@
 Hydraulic conductivity was not varied between layers in the multi-layer models evaluated.
-For models having less than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was averaged for each base model cell contained in a coarse model cell.
-For models having more than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data was directly assigned to each higher-resolution cell contained within a coarse base model cell.
+For models having less than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data were averaged for each base model cell contained in a coarse model cell.
+For models having more than 1,000 $\times$ 1,000 cells per layer, the base hydraulic conductivity data were directly assigned to each higher-resolution cell contained within a coarse base model cell.
 
-Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no flow boundaries at the top and bottom.
+Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no-flow boundaries at the top and bottom.
 Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10.
@@
 = 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively.
-For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively.
+For the one layer 1,000 $\times$ 1,000 cell model, conductances and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively.
 The head-dependent boundaries cause an ambient groundwater flow from left to right.
@@
 Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
-\color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$RCLOSE m.\color{black}
+\color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.\color{black}
 
@@
 The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners.
-An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library.
+An unstructured compressed sparse row (CSR) storage format was used in the UPCG solver to facilitate use of the CUBLAS library.
 The UPCG solver also includes options for running sequential and parallel simulations on the CPU.
@@
 
-A number of test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW.
-CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared to CPU runtimes for the same preconditioner.
+Several test problems were evaluated and indicate that it is possible to realize notable speedup through use of GPGPUs with MODFLOW.
+CPU/GPGPU speedups ranging from 1.5 to 35 were observed for the Jacobi, MILU0, and GLSPOLY preconditioners when GPGPU runtimes were compared with CPU runtimes for the same preconditioner.
 Exceptions to notable CPU/GPGPU speedups were observed for the smallest problems where CPU-GPGPU memory transfer times were significant compared to solution time.
@@
 Source code for MODFLOW (MODFLOW--2005 version) with the UPCG solver is available at \url{https://github.com/jdhughes-usgs/modflow-2005-upcg}.
-Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit which includes the necessary CUDA drivers.
+Development and compilation of CUDA and CUBLAS functions requires installation of the CUDA toolkit, which includes the necessary CUDA drivers.
 NVIDIA also provides GPGPU code samples in the CUDA software development kit (SDK).
@@
 
-The USGS provides no warranty, expressed or implied, as to the correctness of the furnished software or the suitability for any purpose.
-The software has been tested, but as with any complex software, there could be undetected errors.
-Users who find errors are requested to report them to the USGS.
-References to non-USGS products, trade names, and (or) services are provided for information purposes only and do not constitute endorsement or warranty, express or implied, by the USGS, U.S.
-Department of Interior, or U.S.
-Government, as to their suitability, content, usefulness, functioning, completeness, or accuracy.\color{black}
+Although this software has been used by the U.S.
+Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S.
+Government as to the accuracy and functioning of the software program and related program material nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.
+Although the software has been tested, there could be undetected errors.
+Users are encouraged to report any errors to these authors.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
+Government.\color{black}
 
@@
 
-To use GPGPUs, specialized software must be developed to use a precompiled set of functions and routines which serve as a communication mechanism between a CPU and one or more GPGPU and pass instructions to the GPGPU.
+To use GPGPUs, specialized software must be developed to use a precompiled set of functions and routines that serve as a communication mechanism between a CPU and one or more GPGPU and pass instructions to the GPGPU.
 These precompiled functions and routines are typically bundled into a code library that is accessed using a application programming interface (API).
 
-\hspace{15pt}Kernels are blocks of code which are executed only on the GPGPU.
+\hspace{15pt}Kernels are blocks of code that are executed only on the GPGPU.
 When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions.
@@
 \hspace{15pt}Each kernel launches a CUDA grid, which is composed of one or more blocks of grouped threads.
-Threads within the same block can share certain types of physical memory on the GPGPU.
+Threads within the same block share physical memory on the GPGPU.
 The number of blocks in a grid is determined by the size of the kernel data being processed or the number of SM.
@@
 Steady-state groundwater head contours (1 m contour interval) are also shown.
-Note the effect that groundwater withdrawals and hydraulic conductivity are having on simulated groundwater heads.}
+Note the effect that groundwater withdrawals and hydraulic conductivity have on simulated groundwater heads.}
 	\label{FigModelDomain}

### 2012-10-24 GW20120518-0099-r02.tex
revision 02

--- 2012-09-22 GW20120518-0099-r01.tex
@@
 \doublespacing
-\linenumbers
+%\linenumbers
 \allsectionsfont{\sffamily\mdseries\upshape} 
@@
 \renewcommand{\headrulewidth}{0pt}
-\lhead{}\chead{Hughes and White -- MODFLOW UPCG GPGPU Solver}\rhead{}
+%\lhead{}\chead{Hughes and White -- MODFLOW UPCG GPGPU Solver}\rhead{}
+\lhead{}\chead{}\rhead{}
 \lfoot{}\cfoot{\thepage}\rfoot{}
@@
 To evaluate the use of general-purpose graphics processing units (GPGPU) to improve the performance of MODFLOW, an unstructured preconditioned conjugate gradient (UPCG) solver has been developed.
-The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete \color{cyan}lower-upper (LU) \color{black}factorization, and generalized least-squares polynomial preconditioners.
+The UPCG solver uses a compressed sparse row storage scheme and includes Jacobi, zero fill-in incomplete and modified-incomplete lower-upper (LU) factorization, and generalized least-squares polynomial preconditioners.
 The UPCG solver also includes options for sequential and parallel solution on the central processing unit (CPU) using OpenMP.
@@
 The efficiency of the UPCG solver for GPGPU and CPU solutions is benchmarked using simulations of a synthetic, heterogeneous  unconfined aquifer with tens of thousands to millions of active grid cells.
-Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW \color{cyan}preconditioned conjugate gradient (PCG) \color{black}solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
-Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be \color{blue}easily \color{black}parallelized.
+Testing indicates GPGPU speedups on the order of 2-8, relative to the standard MODFLOW preconditioned conjugate gradient (PCG) solver, can be achieved when (1) memory copies between the CPU and GPGPU are optimized, (2) the percentage of time performing memory copies between the CPU and GPGPU is small relative to the calculation time, (3) high-performance GPGPU cards are utilized, and (4) CPU-GPGPU combinations are used to execute sequential operations that are difficult to parallelize.
+Furthermore, UPCG solver testing indicates GPGPU speedups exceed parallel CPU speedups achieved using OpenMP on multi-core CPUs for preconditioners that can be easily parallelized.
 \end{abstract}
@@
 
-MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to \color{blue}numerous \color{black}two- and three-dimensional problems.
+MODFLOW (Harbaugh 2005) is a finite-difference groundwater flow model that has been effectively applied to numerous two- and three-dimensional problems.
 Because MODFLOW uses a cell-centered finite-difference (CCFD) approximation and a rectilinear grid, model runtimes are often increased by factors of two, four, and eight with increased vertical, horizontal, or combined vertical and horizontal discretization, respectively.
-\color{blue}Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
-\color{black}To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
-The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (\color{cyan}SSOR\color{black}) linear solver.
-These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), \color{cyan}the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), \color{black}the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
+Increased model runtimes are typically a direct result of an increase in the number of simultaneous equations solved for each cell but can also be affected (typically to a lesser degree) by changes in the coefficient matrix.
+To date, larger model sizes have been accommodated through a combination of faster CPUs and better linear solvers.
+The original version of MODFLOW (McDonald and Harbaugh 1988) included 1) a strongly implicit procedure (SIP) linear solver and 2) a slice-successive overrelaxation (SSOR) linear solver.
+These original linear solvers were supplemented with the preconditioned conjugate gradient (PCG) linear solver (Hill 1990), the alternating diagonal ordering (DE4) direct solver (Harbaugh 1995), the link-algebraic multi-grid (LMG) linear solver (Mehl and Hill 2001), the geometric multigrid (GMG) linear solver (Wilson and Naff 2004), and the  preconditioned conjugate gradient with improved nonlinear control (PCGN) linear solver (Naff and Banta 2008).
 
 Parallelization of linear solvers is another approach for accommodating larger model sizes.
-\color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem.
+Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem.
 For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem, the speedup would be 2, 3, 4, and 5, respectively.
 However, ideal parallelization is rarely, if ever, achieved because of communication overhead, which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize.
-\color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}
+In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).
 
-\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs \color{cyan}using 8 threads\color{black}.
-Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 \color{cyan}using 32 CPUs\color{black}.
-In addition to OpenMP and MPI, general-purpose graphics processing units (GPGPUs) also \color{blue}present \color{black}an option for linear solver parallelization.
-In comparison to CPUs, GPGPUs \color{blue}typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second.
-\color{black}Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the Compute Unified Device Architecture (CUDA) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
+Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the open multiprocessing (OpenMP) programming paradigm (OpenMP Architecture Review Board, 2005) and observed speedups as large as 5 on machines with multi-core CPUs using 8 threads.
+Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7 using 32 CPUs.
+In addition to OpenMP and MPI, general-purpose graphics processing units (GPGPUs) also present an option for linear solver parallelization.
+In comparison to CPUs, GPGPUs typically have more cores (\textit{e.g.}, 448 cores in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs), have higher bandwidth, and therefore, are capable of executing more floating point operations per second.
+Use of GPGPUs to parallelize linear solvers for MODFLOW is also attractive because of the availability of the Compute Unified Device Architecture (CUDA) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation.
 For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems.
-For solution of the groundwater flow equation, Ji \textit{et al.} (\color{cyan}2010 and \color{black}2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.
+For solution of the groundwater flow equation, Ji \textit{et al.} (2010 and 2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.
 
 In this paper, we evaluate speedups resulting from GPGPU parallelization of the unstructured preconditioned conjugate gradient (UPCG) linear solver for a unconfined groundwater flow problem.
-We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  \color{cyan}lower-upper (MILU0) factorization \color{black}, and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU.
+We also evaluate the performance of the UPCG linear solver using Jacobi, zero fill-in modified incomplete  lower-upper (MILU0) factorization , and generalized least-squares polynomial (GLSPOLY) preconditioners on the CPU and GPGPU.
 And finally, we evaluate the performance of the UPCG solver on the GPGPU to parallel CPU simulations using the OpenMP parallel programming paradigm.
@@
 For GPGPU code implementations, the concept is similar to OpenMP.
-The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA \color{blue}functions, called kernels (\textbf{Box 1})\color{black}.
-\color{blue}However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
+The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA functions, called kernels (\textbf{Box 1}).
+However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
 The NVIDIA Tesla C2050 used to evaluate the UPCG solver in this study can execute 21,504 concurrent threads.
-\color{black} With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization \color{blue} for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations\color{black}.
+With thread counts on the order of tens of thousands, GPGPUs have great potential for parallelization  for MODFLOW and other numerical simulators that use Krylov subspace methods to solve large systems of simultaneous linear equations.
 
@@
 
-In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the \color{cyan}difference between the left and right side (residual) \color{black}of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless].
-\color{cyan}Readers are referred to Barrett \textit{et al.} (1994) for more information on the PCG method.\color{black}
+In \textbf{algorithm~\ref{pcgalg}}, $\mathbf{r}$ is the difference between the left and right side (residual) of \textbf{equation~\ref{lineareq}} [L$^3$/T], $i$ is the linear iteration index [unitless], $maxinner$ is a user-specified maximum number of iterations to perform [unitless], $\mathbf{M}$ is the preconditioned form of $\mathbf{A}$ [L$^2$/T], $\mathbf{z}$ is the solution resulting from application of $\mathbf{M}$ to $\mathbf{r}$ [L$^3$/T], $\mathbf{p}$ is the orthogonal search direction [L], $\beta$ is a scalar used to determine the next search direction [unitless], $\mathbf{q}$ is the residual change resulting from multiplication of $\mathbf{A}$ and the search direction $\mathbf{p}$ [L$^3$/T], $\alpha$ is the step-size [unitless].
+Readers are referred to Barrett \textit{et al.} (1994) for more information on the PCG method.
 
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize \color{cyan}because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).\color{black}
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).
 
 \subsection*{Numerical Implementation}
-The \color{cyan}unstructured preconditioned conjugate gradient (UPCG) \color{black}linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations.
+The unstructured preconditioned conjugate gradient (UPCG) linear solver was coded to allow solution of the groundwater flow equation on either the CPU, GPGPU, or combination of CPU and GPGPU using double precision operations.
 Execution of the UPCG solver on the CPU, GPGPU, or combination of CPU and GPGPU is specified by the user in the UPCG input file.
@@
 Although the connectivity of the finite-difference discretization used in MODFLOW uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row (CSR) storage format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row.
-This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012).
-BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979)\color{black}.
-The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (\color{cyan}Saad \color{black}2003).
+This storage format is also supported by the CUDA-enabled implementation of the BLAS library (CUBLAS -- NVIDIA 2012).
+BLAS is a library of Basic Linear Algebra Subprograms (Lawson \textit{et al.} 1979).
+The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Saad 2003).
 
@@
 For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and vector operations.
-The OpenMP matrix-vector product was constructed using a level-2 BLAS \color{cyan}general matrix-vector (GEMV) product \color{black}approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
+The OpenMP matrix-vector product was constructed using a level-2 BLAS general matrix-vector (GEMV) product approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
 OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively.
@@
 Picard iteration is implemented such that the $\mathbf{A}$ matrix is formulated using the latest estimate of $\mathbf{h}$, then $\mathbf{h}$ and $\mathbf{r}$ are updated using a linear solver.
-This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of Picard (outer) iterations is exceeded; \color{cyan}the UPCG solver approach for evaluating convergence is identical to the standard MODFLOW PCG solver (Hill 1990).\color{black}
+This process is continued until the maximum change in $h$ and $r$ in any model cell is less than HCLOSE and RCLOSE, respectively, on the first iteration ($i$=1) of the linear solver or the maximum number of Picard (outer) iterations is exceeded; the UPCG solver approach for evaluating convergence is identical to the standard MODFLOW PCG solver (Hill 1990).
 
@@
 The performance of the GPGPU implementation of the UPCG linear solver was evaluated on a NVIDIA Tesla C2050 GPGPU with 3 GB of RAM and capable of 515 GFLOP/sec of double precision processing performance (NVIDIA compute capability 2.0).
-The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel \color{cyan}Xeon \color{black}3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS.
-For comparison with GPGPU results, \color{cyan}serial and \color{black}parallel CPU simulations were executed on dual 4-core Intel \color{cyan}Xeon \color{black}2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).
+The GPGPU was mounted in a 16-pin PCI local bus on a single quad core Intel Xeon 3GHz CPU, capable of executing 4 threads per CPU, with 6 GB of RAM and running a 64-bit version of the Windows 7 Enterprise OS.
+For comparison with GPGPU results, serial and parallel CPU simulations were executed on dual 4-core Intel Xeon 2.4 GHz CPUs, capable of executing 8 parallel threads, with 16 GB of RAM running a 64-bit version of the Windows Server 2008 R2 Standard OS (Service Pack 1).
 
-\color{blue}To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes.
+To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes.
 The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear because of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$.
-Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.\color{black}
+Furthermore, the test problems used realistic initial conditions sufficiently far from final results to start with an incorrect $\mathbf{A}$ matrix to maximize (1) head changes during solution of the problem and (2) the number of Picard iterations required to achieve a solution.
 
 \subsection*{Problem Description}
-GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a \color{cyan}1,000 m $\times$ 1,000 m \color{black} heterogeneous unconfined aquifer.
+GPGPU and parallel CPU results were evaluated using a number of horizontal and vertical model discretizations of a 1,000 m $\times$ 1,000 m  heterogeneous unconfined aquifer.
 The top and bottom of the aquifer are flat and specified to have elevations of 10 and -30 m, respectively.
@@
 For multi-layer models, a constant layer thickness was used for all but the first layer and was calculated by dividing 30 m by the number of layers.
-\color{blue}Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.\color{black}
+Layer 1 also includes the thickness of the aquifer above the arbitrary elevation of 0 m and therefore is 10 m thicker than layer 2.
 
@@
 
-Head-dependent (general head) boundaries were specified on the left and right sides \color{cyan} in all layers \color{black} and no-flow boundaries at the top and bottom.
-Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10.
+Head-dependent (general head) boundaries were specified on the left and right sides  in all layers  and no-flow boundaries at the top and bottom.
+Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain $\left [ 10.
 / 1000.
-= 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively.
+= 0.01 \text{ m/m} \right]$ and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively.
 For the one layer 1,000 $\times$ 1,000 cell model, conductances and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively.
@@
 For multi-layer simulations, the pumping wells withdraw water from the lower model layer.
-There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries \color{cyan}and represents 46\% of the flow entering the model\color{black}.
+There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries and represents 46\% of the flow entering the model.
 Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.
@@
 Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads.
-Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
-\color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.\color{black}
+Simulated heads  that satisfy specified convergence criteria  for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels.
+Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$HCLOSE m.
 
 \subsection*{Results}
-\color{blue}The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and hybrid CPU/GPGPU approaches were identical to the number required on the CPU.
+The number of Picard and inner iterations required to solve the test problems on the CPU are summarized on \textbf{table~\ref{TableCPUITSummary}}; the number of Picard and inner iterations required to solve the test problems on the GPGPU and hybrid CPU/GPGPU approaches were identical to the number required on the CPU.
 The number of Picard and inner iterations required by the MILU0 preconditioner is comparable to the standard MODFLOW PCG solver with the modified incomplete Cholesky (MIC) preconditioner (Hill 1990).
@@
 The number of Picard and inner iterations required by the GLSPOLY preconditioner is generally greater than the MILU0 preconditioner and the PCG solver with the MIC preconditioner but it is notably better the Jacobi preconditioner.
-In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.\color{black}
+In fact, the maximum number of Picard and inner (linear) iterations were exceeded for the 4,000 $\times$ 4,000 $\times$ 1 cell problem simulated using the Jacobi preconditioner, highlighting that the Jacobi preconditioner is not a robust preconditioner for some problems.
 
@@
 For the MILU0 preconditioner, all of the speedup occurs during the conjugate gradient routine in the BLAS operations [matrix-vector multiplication and vector operations (dot-products)], which are carried out on the GPGPU after application of the preconditioner.
-Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases \color{cyan}(\textbf{fig.~\ref{FigBLASOperationResults}a} and \textbf{b})\color{black}.
+Furthermore, because of additional memory transfers between the CPU and GPGPU when using the MILU0 preconditioner, the total time spent applying the preconditioner actually increases (\textbf{fig.~\ref{FigBLASOperationResults}a} and \textbf{b}).
 For the GLSPOLY preconditioner, solution on the GPGPU reduces the time spent completing linear solver operations by more than 95\% and the time spent transferring data between the CPU and GPGPU is negligible.
@@
 PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners range from less than 1.0 to more than 8.0, which are notably less than CPU/GPGPU speedups.
-\color{cyan}Reduced PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners relative to CPU/GPGPU speedups can be attributed to the general efficiency of the MODFLOW PCG solver that is a result of serial coding to reduce solver memory requirements.
-\color{black}The relative increase in PCG(MIC)/GPGPU speedups \color{cyan}observed for larger problems with \color{black}the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).
+Reduced PCG(MIC)/GPGPU speedups for the Jacobi and GLSPOLY preconditioners relative to CPU/GPGPU speedups can be attributed to the general efficiency of the MODFLOW PCG solver that is a result of serial coding to reduce solver memory requirements.
+The relative increase in PCG(MIC)/GPGPU speedups observed for larger problems with the Jacobi and GLSPOLY preconditioners is a result of the reduced efficiency of the PCG solver with the MIC preconditioner as the number of model layers increase (\textbf{fig.~\ref{FigGPUResults}b}).
 
 Parallel CPU simulations were run to compare speedups possible with GPGPUs to speedups that could be realized on multi-core CPUs using the OpenMP programming paradigm.
-\color{blue} The PCG solver parallelized by Dong and Li (2009) using OpenMP was not evaluated in this study because we wanted to directly compare CPU, GPGPU, and parallel CPU runtimes for the same solver.
-\color{black}PCG(MIC)/UPCG-OpenMP speedups for the Jacobi, MILU0, and GLSPOLY preconditioners using 4, 7, and 14 OpenMP threads are summarized on \textbf{table~\ref{TableOpenMPSpeedup}}.
+The PCG solver parallelized by Dong and Li (2009) using OpenMP was not evaluated in this study because we wanted to directly compare CPU, GPGPU, and parallel CPU runtimes for the same solver.
+PCG(MIC)/UPCG-OpenMP speedups for the Jacobi, MILU0, and GLSPOLY preconditioners using 4, 7, and 14 OpenMP threads are summarized on \textbf{table~\ref{TableOpenMPSpeedup}}.
 PCG(MIC)/UPCG-OpenMP speedups for the Jacobi and GLSPOLY preconditioners are notably less than PCG(MIC)/GPGPU speedups and is expected because the number of threads on the GPGPU is orders of magnitude greater than for parallel CPU simulations.
-\color{cyan} In some cases, PCG(MIC)/UPCG-OpenMP speedups are less than 1 as a result of the limited number of threads available and the numerical overhead associated with managing OpenMP threads.
-\color{black}For the MILU0 preconditioner, PCG(MIC)/UPCG-OpenMP speedups (average speedup $=$ 1.539) are comparable to PCG(MIC)/GPGPU speedups (average speedup $=$ 1.447) and there is little benefit to using the GPGPU with the MILU0 preconditioner.
+In some cases, PCG(MIC)/UPCG-OpenMP speedups are less than 1 as a result of the limited number of threads available and the numerical overhead associated with managing OpenMP threads.
+For the MILU0 preconditioner, PCG(MIC)/UPCG-OpenMP speedups (average speedup $=$ 1.539) are comparable to PCG(MIC)/GPGPU speedups (average speedup $=$ 1.447) and there is little benefit to using the GPGPU with the MILU0 preconditioner.
 PCG(MIC)/UPCG-OpenMP speedups for the MILU0 preconditioner with 7 OpenMP threads are shown graphically on \textbf{figure~\ref{FigOpenMPResults}}.
@@
 The relatively small PCG(MIC)/UPCG-OpenMP speedups observed with the UPCG solver compiled using the Intel Fortran compiler is consistent with the observations of Dong and Li (2009).
-\color{blue} We observed similar PCG(MIC)/UPCG-OpenMP speedups with and without optimization of the executable compiled using the Intel Fortran compiler.\color{black}
+We observed similar PCG(MIC)/UPCG-OpenMP speedups with and without optimization of the executable compiled using the Intel Fortran compiler.
 
-\color{blue}Based on our results we believe we can offer the following suggestion regarding use of parallel strategies on GPGPUs and CPUs: (1) in cases where the GLSPOLY preconditioner is adequate, solution on the GPGPU will result in significant speedup over serial and parallel CPU solutions; (2) in some cases the use of the Jacobi preconditioner will result in a GPGPU speedup that exceeds the speedup possible with other preconditioner options but this will be the exception rather than the rule and only applicable to select problems (and likely very simple problems); (3) in cases where a stronger preconditioner than the GLSPOLY preconditioner is required, the parallel CPU version of the MILU0 preconditioner will result in the largest speedup; and (4) in cases with smooth, low-frequency error components a completely different solver, such as the LMG or GMG solver, may be needed (see Mehl and Hill 2001).
+Based on our results we believe we can offer the following suggestion regarding use of parallel strategies on GPGPUs and CPUs: (1) in cases where the GLSPOLY preconditioner is adequate, solution on the GPGPU will result in significant speedup over serial and parallel CPU solutions; (2) in some cases the use of the Jacobi preconditioner will result in a GPGPU speedup that exceeds the speedup possible with other preconditioner options but this will be the exception rather than the rule and only applicable to select problems (and likely very simple problems); (3) in cases where a stronger preconditioner than the GLSPOLY preconditioner is required, the parallel CPU version of the MILU0 preconditioner will result in the largest speedup; and (4) in cases with smooth, low-frequency error components a completely different solver, such as the LMG or GMG solver, may be needed (see Mehl and Hill 2001).
 The effectiveness of a preconditioner will be based on the distribution of the eigenvalues of the preconditioned matrix.
@@
 The effectiveness of the Jacobi preconditioner was diminished (that is, required additional iterations) for hydraulic conductivity fields containing values spanning 5 orders of magnitude (results not shown).
-We were able to satisfy convergence criteria for all of the alternative problems evaluated because of the large number of iterations that were allowed (as many as 50,000 inner iterations).\color{black}
+We were able to satisfy convergence criteria for all of the alternative problems evaluated because of the large number of iterations that were allowed (as many as 50,000 inner iterations).
 
@@
 
-The PCG algorithm was implemented for use on GPGPUs using the NVIDIA \color{blue}CUDA API\color{black}.
+The PCG algorithm was implemented for use on GPGPUs using the NVIDIA CUDA API.
 The availability of this API makes it relatively simple to develop parallel GPGPU code.
@@
 
-The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners.
+The UPCG solver was developed to use  the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners.
 An unstructured compressed sparse row (CSR) storage format was used in the UPCG solver to facilitate use of the CUBLAS library.
@@
 When compared to parallel CPU solutions, speedups for GPGPU solutions using the UPCG solver with the Jacobi and GLSPOLY preconditioners are significantly better.
-\color{blue}The primary reason for the effectiveness of parallel GPGPU solutions is the number of cores available on GPGPUs relative to the number available on CPUs.
-\color{black}For problems where the MILU0 preconditioner is needed there is no real benefit to using the GPGPU because of the required additional CPU-GPGPU memory transfers -- for these cases parallel CPU solution would be adequate.
-\color{blue}Improved approaches for parallelization of incomplete factorization preconditioners is an active area of research (for example see, Li and Saad 2010; Naumov 2011; Hwang 2012).
+The primary reason for the effectiveness of parallel GPGPU solutions is the number of cores available on GPGPUs relative to the number available on CPUs.
+For problems where the MILU0 preconditioner is needed there is no real benefit to using the GPGPU because of the required additional CPU-GPGPU memory transfers -- for these cases parallel CPU solution would be adequate.
+Improved approaches for parallelization of incomplete factorization preconditioners is an active area of research (for example see, Li and Saad 2010; Naumov 2011; Hwang 2012).
 For example, Naumov (2011) was able to achieve an average speedup of 2 using a data-dependency graphing approach.
-Therefore, it is expected that additional research will improve the speedup possible with incomplete factorization preconditioners and ultimately the effectiveness of parallel solvers for MODFLOW and other numerical simulators.\color{black}
+Therefore, it is expected that additional research will improve the speedup possible with incomplete factorization preconditioners and ultimately the effectiveness of parallel solvers for MODFLOW and other numerical simulators.
 
@@
 The benefits of using the GPGPU increases as the problem size increases.
-\color{blue}We suspect there are some problems that will require stronger preconditioners than the Jacobi and GLSPOLY preconditioners.
+We suspect there are some problems that will require stronger preconditioners than the Jacobi and GLSPOLY preconditioners.
 For these problems we would suggest the parallel CPU solution with the MILU0 preconditioner.
-As a result, it seems use of hardware having multi-core CPUs and one or more GPGPUs is the most effective strategy for running the range of possible MODFLOW \color{cyan}[including MODFLOW variants such as MODFLOW-FMP2 (Schmid and Hanson 2009)] \color{blue} simulations in parallel.\color{black}
+As a result, it seems use of hardware having multi-core CPUs and one or more GPGPUs is the most effective strategy for running the range of possible MODFLOW [including MODFLOW variants such as MODFLOW-FMP2 (Schmid and Hanson 2009)]  simulations in parallel.
 
@@
 The UPCG solver is designed to solve  square, symmetric, positive-definite coefficient matrices.
-Modification of the UPCG solver to allow solution of nonsymmetric matrices, using the BiConjugate Gradient Stabilized method or other similar method (Barrett \textit{et al.} 1994), would permit use of the UPCG solver with MODFLOW-NWT (Niswonger \textit{et al.} 2011), MT3DMS (Zheng and Wang 1999), SEAWAT (Langevin \textit{et al.} 2007), \color{cyan} or the SWR1 Process for MODFLOW-2005 and MODFLOW-NWT (Hughes \textit{et al.} 2012)\color{black}.
+Modification of the UPCG solver to allow solution of nonsymmetric matrices, using the BiConjugate Gradient Stabilized method or other similar method (Barrett \textit{et al.} 1994), would permit use of the UPCG solver with MODFLOW-NWT (Niswonger \textit{et al.} 2011), MT3DMS (Zheng and Wang 1999), SEAWAT (Langevin \textit{et al.} 2007),  or the SWR1 Process for MODFLOW-2005 and MODFLOW-NWT (Hughes \textit{et al.} 2012).
 
@@
 
-\color{blue}
 \section* {\bf How to Obtain the Software}
@@
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
-Government.\color{black}
+Government.
 
@@
 
-\item \color{cyan} Harbaugh, A.W.
+\item  Harbaugh, A.W.
 1995.
@@
 Geological Survey modular finite-difference ground-water flow model: U.S.
-Geological Survey Open-File Report 95-288, 46 p.\color{black}
+Geological Survey Open-File Report 95-288, 46 p.
 
@@
 
-\item \color{cyan} Hughes, J.D., Langevin, C.D., Chartier, K.L., White, J.T.
+\item  Hughes, J.D., Langevin, C.D., Chartier, K.L., White, J.T.
 2012.
@@
 Geological Survey Techniques and Methods, book 6, chap.
-A40 (Version 1.0), 113 p.\color{black}
+A40 (Version 1.0), 113 p.
 
-\item \color{blue} Hwang, H.T.
+\item  Hwang, H.T.
 2012.
 Development of a parallel computational framework to solve flow and transport in integrated surface-subsurface hydrologic systems.
-PhD dissertation, University of Waterloo, 133 p.\color{black}
+PhD dissertation, University of Waterloo, 133 p.
 
-\item \color{cyan} Ji, Xiaohui, Cheng, Tangpei, Wang, Qun 2010.
+\item  Ji, Xiaohui, Cheng, Tangpei, Wang, Qun 2010.
 A simulation of large-scale groundwater flow on CUDA-enabled GPUs.
@@
 ACM, New York, NY, USA, 2402-2403.
-doi: 10.1145/1774088.1774588\color{black}
+doi: 10.1145/1774088.1774588
 
@@
 
-\item \color{cyan}Lawson, C.
+\item Lawson, C.
 L., Hanson, R.
@@
 3, 308-323.
-doi: 10.1145/355841.355847\color{black}
+doi: 10.1145/355841.355847
 
@@
 
-\item \color{blue} Naumov, Maxim 2011.
+\item  Naumov, Maxim 2011.
 Incomplete-LU and Cholesky preconditioned iterative methods using CUSPARSE and CUBLAS.
 NVIDIA technical report, 16 p.
-Available from \url{http://developer.download.nvidia.com/devzone//devcenter/cuda/files/psts_white_paper_final.pdf}\color{black}
+Available from \url{http://developer.download.nvidia.com/devzone//devcenter/cuda/files/psts_white_paper_final.pdf}
 
@@
 
-\item \color{cyan} NVIDIA 2012.
+\item  NVIDIA 2012.
 CUDA Toolkit 4.2 -- CUBLAS Library: PG-05326-041\_v01, 99 p.
-Available from \url{http://developer.download.nvidia.com/compute/DevZone/docs/html/CUDALibraries/doc/CUBLAS_Library.pdf}  \color{black}
+Available from \url{http://developer.download.nvidia.com/compute/DevZone/docs/html/CUDALibraries/doc/CUBLAS_Library.pdf}
 
@@
 
-\item \color{cyan}Schmid, Wolfgang and Hanson R.T.
+\item Schmid, Wolfgang and Hanson R.T.
 2009.
 The Farm Process Version 2 (FMP2) for MODFLOW-2005 - Modifications and Upgrades to FMP1: U.S.
-Geological Survey Techniques in Water Resources Investigations, Book 6, Chapter A32, 102p.\color{black}
+Geological Survey Techniques in Water Resources Investigations, Book 6, Chapter A32, 102p.
 
@@
 \parbox[c]{5.75in}{\label{GPGPUbox1}
-\color{blue}\textbf{\\Box 1\\Developing Software for GPGPUs}
+\textbf{\\Box 1\\Developing Software for GPGPUs}
 
@@
 When the code is compiled, the compiler is instructed to use the API functions and routines to resolve the kernel block of code into GPGPU instructions.
-The kernels executed by the UPCG solver on the GPGPU are the CUDA-enabled BLAS operations in the preconditioned conjugate gradient algorithm.\\ \color{black}}}
+The kernels executed by the UPCG solver on the GPGPU are the CUDA-enabled BLAS operations in the preconditioned conjugate gradient algorithm.\\ }}
 \bigskip
@@
 \parbox{5.75in}{
-\color{blue}\textbf{\\Box 2\\NVIDIA GPGPU Architecture and Kernel Execution}
+\textbf{\\Box 2\\NVIDIA GPGPU Architecture and Kernel Execution}
 
@@
 \hspace{15pt}The Tesla C2050 used in this study has the following specifications: SMs = 14, CC per SM = 32, threads per block = 1,024, and threads per warp = 32.
-The multi-threaded instruction unit on each of the 14 Tesla C2050 SMs can run 48 warps concurrently, yielding a total of 21,504 concurrent threads (14 SMs $\times$ 48 warps per SM $\times$ 32 threads per warp).\\ \color{black}
+The multi-threaded instruction unit on each of the 14 Tesla C2050 SMs can run 48 warps concurrently, yielding a total of 21,504 concurrent threads (14 SMs $\times$ 48 warps per SM $\times$ 32 threads per warp).\\
 }}
@@
 
-\begin{figure}[hp]
+\begin{figure}
  	\centering
-  	\includegraphics[width=17.15cm]{Figure1.png}
+  	%\includegraphics[width=17.15cm]{Figure1.png}
 	\caption{(A) Fork-join parallel model.
@@
 %Figure 2
-\newpage
-\begin{figure}[hp]
+\begin{figure}
 	\centering
-  	\includegraphics[width=8.25cm]{Figure2.png}
+  	%\includegraphics[width=8.25cm]{Figure2.png}
  	\caption{Model domain showing the distribution of hydraulic conductivity and boundary conditions applied to a hypothetical unconfined aquifer.
@@
 %Figure 3
-\newpage
-\begin{figure}[hp]
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure3a.png}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure3b.png}
-	\end{subfigure}
+\begin{figure}
+	%\begin{subfigure}[b]{0.5\textwidth}
+	%	\centering
+  	%	\includegraphics[width=8.25cm]{Figure3a.png}
+	%\end{subfigure}
+	%\begin{subfigure}[b]{0.5\textwidth}
+	%	\centering
+  	%	\includegraphics[width=8.25cm]{Figure3b.png}
+	%\end{subfigure}
  	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to sequential UPCG solver simulations executed on the CPU.
@@
 %Figure 4
-\newpage
-\begin{figure}[hp]
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure4a.png}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure4b.png}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure4c.png}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure4d.png}
-	\end{subfigure}
+\begin{figure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure4a.png}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure4b.png}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure4c.png}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure4d.png}
+%	\end{subfigure}
  	\caption{Time, in seconds, spent applying the preconditioner, calculating matrix-vector products, calculating \texttt{axpy}, calculating all other vector operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
@@
 %Figure 5
-\newpage
-\begin{figure}[hp]
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure5a.png}
-	\end{subfigure}
-	\begin{subfigure}[b]{0.5\textwidth}
-		\centering
-  		\includegraphics[width=8.25cm]{Figure5b.png}
-	\end{subfigure}
+\begin{figure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure5a.png}
+%	\end{subfigure}
+%	\begin{subfigure}[b]{0.5\textwidth}
+%		\centering
+%  		\includegraphics[width=8.25cm]{Figure5b.png}
+%	\end{subfigure}
  	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}
@@
 %Figure 6
-\newpage
-\begin{figure}[hp]
+\begin{figure}
 	\centering
-  	\includegraphics[width=8.25cm]{Figure6.png}
+%  	\includegraphics[width=8.25cm]{Figure6.png}
  	\caption{Speedup of parallel CPU simulations using the UPCG solver with the MILU0 preconditioner and 7 threads relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}

### 2012-10-27 GW20120518-0099-r02.tex
Revision 2 submitted to journal

@@
 For GPGPU code implementations, the concept is similar to OpenMP.
-The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA functions, called kernels (\textbf{Box 1}).
+The main code is a sequential host code, which forks the data into many threads at the invocation of a CUDA function, called a kernel (\textbf{Box 1}).
 However, as opposed to the few threads spawned by OpenMP, GPGPUs have the potential to spawn thousands to tens of thousands of threads to execute the CUDA kernel function (\textbf{Box 2}).
@@
 Application of the incomplete factorization preconditioners requires a forward substitution ($\mathbf{Ly = b}$ -- where $\mathbf{L}$ is the lower triangular portion of $\mathbf{A}$) followed by a backward substitution ($\mathbf{Uz = y}$ – where $\mathbf{U}$ is upper triangular portion of $\mathbf{A}$) and is difficult to efficiently parallelize (Barrett \textit{et al.} 1994).
-Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector multiplied by a constant to another vector (\texttt{axpy}).
+Preconditioning using polynomial preconditioners, however, is simple, efficient, and easy to parallelize because polynomial preconditioner application only requires a limited number of operations (Liang \textit{et al.} 2002; Saad 2003); the operations required to apply a polynomial preconditioner in \textbf{algorithm~\ref{pcgalg}} include (1) matrix-vector products and (2) addition of a vector-scalar product to another vector (referred to as \texttt{axpy}, which is short for \textbf{a}lpha times \textbf{x} \textbf{p}lus \textbf{y}).
 
@@
 For parallel CPU simulations, OpenMP directives were added to all matrix-vector products and vector operations.
-The OpenMP matrix-vector product was constructed using a level-2 BLAS general matrix-vector (GEMV) product approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
+The OpenMP matrix-vector product was constructed using a level-2 BLAS general matrix-vector (\texttt{GEMV}) product approach because of the slight increase in performance of this approach over dot-product and \texttt{axpy} approaches for calculating matrix vector products (Petersen and Arbenz 2004).
 OpenMP directives \texttt{NUM\_THREADS} and \texttt{SCHEDULE STATIC} were used in all OpenMP routines to allow the user to control the number of OpenMP threads and reduce OpenMP scheduling/synchronization overhead, respectively.
@@
 
-To evaluate the performance of the GPGPU implementation of the UPCG linear solver a  synthetic test problem with aquifer properties and boundary conditions that result in equivalent problems and scalable runtimes when discretized using different horizontal and vertical cell sizes.
+To evaluate the performance of the GPGPU implementation of the UPCG linear solver we constructed a synthetic test problem having aquifer properties and boundary conditions that could be scaled to models with different horizontal and vertical cell sizes but still produce equivalent results and scalable runtimes.
 The test problem represents an unconfined aquifer to increase the difficulty of the problem by making it non-linear because of the dependence of the $\mathbf{A}$ matrix in \textbf{equation~\ref{lineareq}} on the solution of $\mathbf{h}$.
@@
  	\centering
-  	%\includegraphics[width=17.15cm]{Figure1.png}
 	\caption{(A) Fork-join parallel model.
@@
 	\centering
-  	%\includegraphics[width=8.25cm]{Figure2.png}
  	\caption{Model domain showing the distribution of hydraulic conductivity and boundary conditions applied to a hypothetical unconfined aquifer.
@@
 \begin{figure}
-	%\begin{subfigure}[b]{0.5\textwidth}
-	%	\centering
-  	%	\includegraphics[width=8.25cm]{Figure3a.png}
-	%\end{subfigure}
-	%\begin{subfigure}[b]{0.5\textwidth}
-	%	\centering
-  	%	\includegraphics[width=8.25cm]{Figure3b.png}
-	%\end{subfigure}
  	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to sequential UPCG solver simulations executed on the CPU.
@@
 \begin{figure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure4a.png}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure4b.png}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure4c.png}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure4d.png}
-%	\end{subfigure}
  	\caption{Time, in seconds, spent applying the preconditioner, calculating matrix-vector products, calculating \texttt{axpy}, calculating all other vector operations, and transferring data between the CPU and GPGPU for (A) 2,000 $\times$ 2,000 $\times$ 3 problem using the MILU0 preconditioner, (B) 4,000 $\times$ 4,000 $\times$ 1 problem using the MILU0 preconditioner, (C) 2,000 $\times$ 2,000 $\times$ 3 problem using the GLSPOLY preconditioner, and (D) 4,000 $\times$ 4,000 $\times$ 1 problem using the GLSPOLY preconditioner.
@@
 \begin{figure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure5a.png}
-%	\end{subfigure}
-%	\begin{subfigure}[b]{0.5\textwidth}
-%		\centering
-%  		\includegraphics[width=8.25cm]{Figure5b.png}
-%	\end{subfigure}
  	\caption{Speedup of UPCG solver GPGPU simulations with the (A) MILU0 and (B) GLSPOLY preconditioners relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}
@@
 	\centering
-%  	\includegraphics[width=8.25cm]{Figure6.png}
  	\caption{Speedup of parallel CPU simulations using the UPCG solver with the MILU0 preconditioner and 7 threads relative to simulations performed using the PCG solver with the modified incomplete Cholesky (MIC) preconditioner.}
